---
title: "Prophet fDOM Forecasting — Streamlined Workflow"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5
)
```

---

## Packages and Data

Uses Meta's `prophet` package. Prophet decomposes the time series into trend +
weekly + yearly seasonality components. External covariates are added as additive
regressors (Melofton et al. approach).

Prophet is the **cleanest fit-once model** in this workflow — no state update or
`refit()` is needed. `predict(model, future_df)` accepts any set of future dates
directly, so forecasting from any 2024 reference date is just one `predict()` call
with the 30 future dates.

Prophet requires the target column to be named `y` and the date column `ds`.
These are renamed internally — the rest of the script uses the same conventions
as all other models.

```{r packages}
library(tidyverse)
library(lubridate)
library(prophet)
library(zoo)        # na.approx for covariate gap-filling
```

```{r load-data}
getwd()
ar_df <- read_csv("../Data/Daily_catwalk_met_RH_2021_2025_lagZS.csv") |>
  mutate(Date = as.Date(Date)) |>
  select(Date, fDOM_1_QSU_daily,
         Temp_1_C_daily, DOsat_1_pct_daily, Chla_1_ugL_daily,
         RH_Q_m3day, RH_DOC_mgL)

# Training window: all data before 2024
train_cutoff <- ymd("2024-01-01")

# Prediction window: every day of 2024
pred_dates_2024 <- seq(ymd("2024-01-01"), ymd("2024-12-31"), by = "day")
```

---

## Helper: Interpolate NAs

Used to fill covariate gaps before fitting and before forecast-step covariate lookup.

```{r fn-interpolate}
interpolate_ts <- function(x) {
  if (all(is.na(x))) return(x)
  zoo::na.approx(x, na.rm = FALSE, rule = 2)
}
```

---

## Model Fitting

### Fitting function

Fits a Prophet model on data before `train_cutoff` following Melofton et al.:

- Linear growth
- Yearly + weekly seasonality (additive)
- No daily seasonality
- External covariates as additive regressors (when supplied)
- `mcmc.samples = 0`, `uncertainty.samples = 0` — point estimates only (fast)

Returns the fitted model, a training plot, a components decomposition plot
(trend + seasonality), and metadata for forecasting.

```{r fn-fit}
fit_prophet_model <- function(all_data,
                               train_cutoff,
                               target,
                               covariates  = NULL,
                               model_name,
                               seed        = 100) {

  set.seed(seed)

  col_select <- c(target, covariates)

  # Build training data — interpolate covariate NAs, then drop remaining NAs
  # Prophet requires ds (date) and y (target) column names
  train_df <- all_data |>
    filter(Date < train_cutoff) |>
    select(Date, all_of(col_select)) |>
    arrange(Date) |>
    mutate(across(-Date, interpolate_ts)) |>
    na.omit() |>
    rename(ds = Date, y = !!sym(target))

  message(model_name, " | n_train = ", nrow(train_df))

  # Initialize Prophet model (Melofton et al. specification)
  m <- prophet(
    growth               = "linear",
    fit                  = FALSE,
    yearly.seasonality   = TRUE,
    weekly.seasonality   = TRUE,
    daily.seasonality    = FALSE,
    seasonality.mode     = "additive",
    mcmc.samples         = 0,
    uncertainty.samples  = 0
  )

  # Add external regressors (additive mode, auto-standardized)
  if (!is.null(covariates)) {
    for (cv in covariates) {
      m <- add_regressor(m, name = cv, standardize = "auto", mode = "additive")
    }
  }

  # Fit model
  m <- fit.prophet(m = m, df = train_df)

  # In-sample fitted values — predict() on training dates (drop y, keep ds + covariates)
  train_pred_df <- train_df |> select(-y)
  fitted_vals   <- predict(m, train_pred_df)

  train_df_plot <- train_df |>
    mutate(fitted = fitted_vals$yhat)

  train_rmse <- sqrt(mean((train_df_plot$y - train_df_plot$fitted)^2, na.rm = TRUE))

  # Training plot: observed vs fitted timeseries
  p_fit <- ggplot(train_df_plot) +
    geom_point(aes(x = ds, y = y, color = "Observed"),
               size = 1.2, alpha = 0.8) +
    geom_point(aes(x = ds, y = fitted, color = "Prophet fitted"),
               size = 1.2, alpha = 0.8) +
    scale_color_manual(values = c("Observed" = "black", "Prophet fitted" = "dodgerblue"),
                       breaks = c("Observed", "Prophet fitted")) +
    labs(title = paste0(model_name, " — training fit (pre-", year(train_cutoff), ")",
                        "  |  RMSE = ", round(train_rmse, 2), " QSU"),
         x = "Date", y = "fDOM (QSU)", color = NULL) +
    theme_bw() +
    theme(legend.position = "top")

  # Components plot: trend + weekly + yearly seasonality decomposition
  p_components <- prophet_plot_components(m, fitted_vals)

  list(
    model            = m,
    plot             = p_fit,
    components_plot  = p_components,
    model_name       = model_name,
    target           = target,
    covariates       = covariates,
    train_rmse       = train_rmse
  )
}
```

### Fit 4 models

All models trained on data before `r train_cutoff`. Prophet's trend and seasonality
parameters are fixed here and reused for all 2024 forecasts — no state update needed.

```{r fit-models}
# Prophet1: trend + seasonality only (no external regressors)
Prophet_lagonly <- fit_prophet_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  covariates   = NULL,
  model_name   = "Prophet_lag_only"
)

# Prophet2: + in-lake (Temp, DO, Chla)
Prophet_lake <- fit_prophet_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  covariates   = c("Temp_1_C_daily", "DOsat_1_pct_daily", "Chla_1_ugL_daily"),
  model_name   = "Prophet_lag_lake"
)

# Prophet3: + catchment (RH discharge and DOC)
Prophet_catch <- fit_prophet_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  covariates   = c("RH_Q_m3day", "RH_DOC_mgL"),
  model_name   = "Prophet_lag_catchment"
)

# Prophet4: + in-lake + catchment
Prophet_lake_catch <- fit_prophet_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  covariates   = c("Temp_1_C_daily", "DOsat_1_pct_daily", "Chla_1_ugL_daily",
                   "RH_Q_m3day", "RH_DOC_mgL"),
  model_name   = "Prophet_lag_lake_catchment"
)
```

### Training plots

```{r plot-training, fig.height=4}
Prophet_lagonly$plot
Prophet_lake$plot
Prophet_catch$plot
Prophet_lake_catch$plot
```

### Components plots

Trend + weekly + yearly seasonality decomposition for each model.

```{r plot-components, fig.height=6}
Prophet_lagonly$components_plot
Prophet_lake$components_plot
Prophet_catch$components_plot
Prophet_lake_catch$components_plot
```

---

## 30-Day Forecast Function

Prophet is the simplest forecast function in this workflow — no for-loop, no state
update, no `refit()`. `predict(model, future_df)` accepts any future dates directly.

For covariate models, observed covariate values for the 30 forecast days are joined
to the future dataframe (perfect covariate assumption, same as all other scripts).
The `yhat` column from `predict()` is returned as `fDOM_1m_predicted`.

```{r fn-forecast}
# # inputs to test function
# forecast_date <- ymd("2024-01-01")
# model_obj <- Prophet_lake_catch
# all_data <- ar_df
# target  = "fDOM_1_QSU_daily"
# horizon = 30

forecast_fdom_30day <- function(forecast_date,
                                 model_obj,
                                 all_data,
                                 target  = "fDOM_1_QSU_daily",
                                 horizon = 30) {

  covariates <- model_obj$covariates

  # Check observed fDOM exists on reference date
  curr_fdom <- all_data |>
    filter(Date == forecast_date) |>
    pull(!!sym(target))

  if (length(curr_fdom) == 0 || is.na(curr_fdom)) return(NULL)

  # Forecast dates: day 1 through day 30 after reference date
  forecast_dates <- seq(forecast_date + 1, forecast_date + horizon, by = "day")

  # Build future dataframe — Prophet requires ds column
  future_df <- tibble(ds = forecast_dates)

  # Add observed covariate values for forecast period (perfect covariate assumption)
  if (!is.null(covariates)) {
    covar_future <- all_data |>
      filter(Date %in% forecast_dates) |>
      select(Date, all_of(covariates)) |>
      arrange(Date) |>
      tidyr::complete(Date = forecast_dates) |>
      mutate(across(-Date, interpolate_ts)) |>
      rename(ds = Date)

    future_df <- future_df |>
      left_join(covar_future, by = "ds")

    if (any(is.na(future_df |> select(-ds)))) {
      future_df <- future_df |>
        mutate(across(-ds, interpolate_ts))
    }
  }

  # Generate forecast — single predict() call, no looping needed
  fc <- predict(model_obj$model, future_df)

  tibble(Date = as.Date(fc$ds), fDOM_1m_predicted = fc$yhat) |>
    mutate(
      model_id       = model_obj$model_name,
      reference_date = forecast_date,
      Horizon        = as.integer(Date - forecast_date)
    ) |>
    select(model_id, reference_date, forecast_date = Date, Horizon, fDOM_1m_predicted) |>
    filter(Horizon > 0)
}

# # test function
# zz <- forecast_fdom_30day(forecast_date = ymd("2024-01-01"), model_obj = Prophet_lake_catch,
#                     all_data = ar_df, target = "fDOM_1_QSU_daily", horizon = 30)
```

---

## Run Forecasts Across 2024

```{r fn-loop}
run_forecasts <- function(model_obj,
                           all_data,
                           pred_dates,
                           target  = "fDOM_1_QSU_daily",
                           horizon = 30) {

  valid_dates <- all_data |>
    filter(Date %in% pred_dates, !is.na(!!sym(target))) |>
    pull(Date)

  message(model_obj$model_name, " | forecasting ", length(valid_dates),
          " reference dates...")

  map_dfr(valid_dates, \(d) forecast_fdom_30day(
    forecast_date = d,
    model_obj     = model_obj,
    all_data      = all_data,
    target        = target,
    horizon       = horizon
  ))
}
```

```{r run-forecasts}
forecasts_prophet1 <- run_forecasts(Prophet_lagonly,    ar_df, pred_dates_2024)
forecasts_prophet2 <- run_forecasts(Prophet_lake,       ar_df, pred_dates_2024)
forecasts_prophet3 <- run_forecasts(Prophet_catch,      ar_df, pred_dates_2024)
forecasts_prophet4 <- run_forecasts(Prophet_lake_catch, ar_df, pred_dates_2024)

all_Prophet_forecasts <- bind_rows(forecasts_prophet1, forecasts_prophet2,
                                    forecasts_prophet3, forecasts_prophet4)

glimpse(all_Prophet_forecasts)
```

---

## Save Output

```{r save}
write_csv(all_Prophet_forecasts,
          "../Data/Prophet_forecasts_2024_25may26.csv")


```
