---
title: "ETS fDOM Forecasting — Streamlined Workflow"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5
)
```

---

## Packages and Data

Uses `fable::ETS()` (Error, Trend, Seasonality exponential smoothing). ETS does not
support external covariates, so the 4-model structure here reflects different ETS
**specifications** rather than covariate sets (following Melofton et al.):

| Model | Specification | Description |
|---|---|---|
| ETS_auto | `ETS(y)` | Fully automatic selection (AIC) |
| ETS_SES | `error("A")` | Simple Exponential Smoothing |
| ETS_Holt | `error("A") + trend("A")` | Holt's linear trend |
| ETS_damped | `error("A") + trend("Ad")` | Damped trend |

Forecasting uses `fabletools::refit(..., reestimate = FALSE)` to update the smoothing
state (level, trend) to the reference date without changing the smoothing parameters —
giving the same fit-once structure as all other scripts.

```{r packages}
library(tidyverse)
library(lubridate)
library(fable)      # ETS()
library(fabletools) # refit()
library(tsibble)
library(zoo)        # na.approx
```

```{r load-data}
getwd()
ar_df <- read_csv("../Data/Daily_catwalk_met_RH_2021_2025_lagZS.csv") |>
  mutate(Date = as.Date(Date)) |>
  select(Date, fDOM_1_QSU_daily)   # ETS is univariate — no covariates needed

# Training window: all data before 2024
train_cutoff <- ymd("2024-01-01")

# Prediction window: every day of 2024
pred_dates_2024 <- seq(ymd("2024-01-01"), ymd("2024-12-31"), by = "day")
```

---

## Helper: Interpolate NAs

```{r fn-interpolate}
interpolate_ts <- function(x) {
  if (all(is.na(x))) return(x)
  zoo::na.approx(x, na.rm = FALSE, rule = 2)
}
```

---

## Model Fitting

### Fitting function

`ets_spec`: right-hand side of the ETS formula as a string
(e.g. `"error('A') + trend('Ad') + season('N')"`) or `NULL` for fully automatic selection.

Season is fixed to `"N"` (none) for all explicit specs — daily water quality data
at sub-annual scale does not warrant a seasonal ETS component here.

Returns the fitted mable, training plot, and metadata.

```{r fn-fit}
fit_ets_model <- function(all_data,
                           train_cutoff,
                           target,
                           ets_spec   = NULL,
                           model_name,
                           seed       = 100) {

  set.seed(seed)

  train_df <- all_data |>
    filter(Date < train_cutoff) |>
    select(Date, all_of(target)) |>
    arrange(Date) |>
    tidyr::complete(Date = seq(min(Date), max(Date), by = "day")) |>
    mutate(across(-Date, interpolate_ts))

  train_ts <- as_tsibble(train_df, index = Date)

  message(model_name, " | n_train = ", nrow(train_df))

  # Fit ETS — NULL spec uses full auto-selection; otherwise RHS spec is injected
  if (is.null(ets_spec)) {
    ets_fit <- train_ts |>
      model(ets = fable::ETS(!!sym(target)))
  } else {
    formula_obj <- as.formula(paste(target, "~", ets_spec))
    ets_fit <- train_ts |>
      model(ets = fable::ETS(formula = formula_obj))
  }

  # Report selected ETS components
  message(model_name, " | ", capture.output(report(ets_fit))[1])

  # In-sample fitted values
  fitted_vals <- fitted(ets_fit) |>
    as_tibble() |>
    select(Date, .fitted)

  train_df_plot <- train_df |>
    left_join(fitted_vals, by = "Date")

  train_rmse <- sqrt(mean((train_df_plot[[target]] - train_df_plot$.fitted)^2,
                           na.rm = TRUE))

  # Training plot: observed vs fitted timeseries
  p_fit <- ggplot(train_df_plot) +
    geom_point(aes(x = Date, y = !!sym(target), color = "Observed"),
               size = 1.2, alpha = 0.8) +
    geom_point(aes(x = Date, y = .fitted, color = "ETS fitted"),
               size = 1.2, alpha = 0.8) +
    scale_color_manual(values = c("Observed" = "black", "ETS fitted" = "dodgerblue"),
                       breaks = c("Observed", "ETS fitted")) +
    labs(title = paste0(model_name, " — training fit (pre-", year(train_cutoff), ")",
                        "  |  RMSE = ", round(train_rmse, 2), " QSU"),
         x = "Date", y = "fDOM (QSU)", color = NULL) +
    theme_bw() +
    theme(legend.position = "top")

  list(
    model      = ets_fit,
    plot       = p_fit,
    model_name = model_name,
    target     = target,
    covariates = NULL,       # ETS is always univariate
    train_rmse = train_rmse
  )
}
```

### Fit 4 models

```{r fit-models}
# ETS1: fully automatic selection
ETS_auto <- fit_ets_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  ets_spec     = NULL,
  model_name   = "ETS_auto"
)

# ETS2: Simple Exponential Smoothing (level only, additive error)
ETS_SES <- fit_ets_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  ets_spec     = "error('A') + trend('N') + season('N')",
  model_name   = "ETS_SES"
)

# ETS3: Holt's linear trend (additive error + additive trend)
ETS_Holt <- fit_ets_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  ets_spec     = "error('A') + trend('A') + season('N')",
  model_name   = "ETS_Holt"
)

# ETS4: Damped trend (Melofton et al. approach)
ETS_damped <- fit_ets_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  ets_spec     = "error('A') + trend('Ad') + season('N')",
  model_name   = "ETS_damped"
)
```

### Training plots

```{r plot-training, fig.height=4}
ETS_auto$plot
ETS_SES$plot
ETS_Holt$plot
ETS_damped$plot
```

---

## 30-Day Forecast Function

`refit(..., reestimate = FALSE)` runs the observed history through the fixed ETS
equations to update the state vector (level, trend) to the reference date. Smoothing
parameters (α, β, φ) are unchanged from training.

`forecast(h = 30)` then projects the updated state forward 30 days.

```{r fn-forecast}
# # inputs to test function
# forecast_date <- ymd("2024-01-01")
# model_obj <- ETS_damped
# all_data <- ar_df
# target  = "fDOM_1_QSU_daily"
# horizon = 30

forecast_fdom_30day <- function(forecast_date,
                                 model_obj,
                                 all_data,
                                 target  = "fDOM_1_QSU_daily",
                                 horizon = 30) {

  # Check observed fDOM exists on reference date
  curr_fdom <- all_data |>
    filter(Date == forecast_date) |>
    pull(!!sym(target))

  if (length(curr_fdom) == 0 || is.na(curr_fdom)) return(NULL)

  # Full observed history up to and including reference date (for state update)
  history_ts <- all_data |>
    filter(Date <= forecast_date) |>
    select(Date, all_of(target)) |>
    arrange(Date) |>
    tidyr::complete(Date = seq(min(Date), max(Date), by = "day")) |>
    mutate(across(-Date, interpolate_ts)) |>
    as_tsibble(index = Date)

  # Update ETS state with fixed smoothing parameters (no re-estimation)
  model_updated <- model_obj$model |>
    refit(new_data = history_ts, reestimate = FALSE)

  # Forecast 30 days ahead from updated state
  fc <- forecast(model_updated, h = horizon)

  # Extract mean forecast and format to match all other model outputs
  fc |>
    as_tibble() |>
    select(Date, .mean) |>
    mutate(
      model_id          = model_obj$model_name,
      reference_date    = forecast_date,
      Horizon           = as.integer(Date - forecast_date),
      fDOM_1m_predicted = .mean
    ) |>
    select(model_id, reference_date, forecast_date = Date, Horizon, fDOM_1m_predicted) |>
    filter(Horizon > 0)
}

# # test function
# zz <- forecast_fdom_30day(forecast_date = ymd("2024-01-01"), model_obj = ETS_damped,
#                     all_data = ar_df, target = "fDOM_1_QSU_daily", horizon = 30)
```

---

## Run Forecasts Across 2024

```{r fn-loop}
run_forecasts <- function(model_obj,
                           all_data,
                           pred_dates,
                           target  = "fDOM_1_QSU_daily",
                           horizon = 30) {

  valid_dates <- all_data |>
    filter(Date %in% pred_dates, !is.na(!!sym(target))) |>
    pull(Date)

  message(model_obj$model_name, " | forecasting ", length(valid_dates),
          " reference dates...")

  map_dfr(valid_dates, \(d) forecast_fdom_30day(
    forecast_date = d,
    model_obj     = model_obj,
    all_data      = all_data,
    target        = target,
    horizon       = horizon
  ))
}
```

```{r run-forecasts}
forecasts_ets1 <- run_forecasts(ETS_auto,   ar_df, pred_dates_2024)
forecasts_ets2 <- run_forecasts(ETS_SES,    ar_df, pred_dates_2024)
forecasts_ets3 <- run_forecasts(ETS_Holt,   ar_df, pred_dates_2024)
forecasts_ets4 <- run_forecasts(ETS_damped, ar_df, pred_dates_2024)

all_ETS_forecasts <- bind_rows(forecasts_ets1, forecasts_ets2,
                                forecasts_ets3, forecasts_ets4)

glimpse(all_ETS_forecasts)
```

---

## Save Output

```{r save}
write_csv(all_ETS_forecasts,
          "../Data/ETS_forecasts_2024.csv")

message("Saved ", nrow(all_ETS_forecasts), " forecast rows across ",
        n_distinct(all_ETS_forecasts$model_id), " models.")
```
