---
title: "Evaluate Predictions"
author: "DWH"
date: "2026-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages and Data

```{r packages}
library(tidyverse)
library(lubridate)
```

```{r load-data}
eval <- read_csv("../Data/Daily_catwalk_met_RH_2021_2025_lagZS.csv") |>
  mutate(Date = as.Date(Date)) |> select(Date, fDOM_1_QSU_daily) |> 
  rename(fDOM_1m_obs = fDOM_1_QSU_daily)


##4 cast outputs 
AR <- read_csv("../Data/AR_forecasts_2024_22may26.csv")

Persistence <- read_csv("../Data/Persistence_forecasts_2024_22may26.csv")

RF <- read_csv("../Data/RF_forecasts_2024_22may26.csv")

NNETAR <- read_csv("../Data/NNETAR_forecasts_2024_22may26.csv")


#join predictiosn 
predictions <- rbind(AR, Persistence, RF, NNETAR)


#### join
Pred_eval_df <- left_join(predictions, eval, by = c("forecast_date" =  "Date" )) 

```

---

## Function Definitions


### RMSE: `calc_rmse_by_horizon()`

Returns RMSE at each forecast horizon. Set `by_season = TRUE` to additionally
group by season (Winter / Spring / Summer / Fall).

```{r fn-rmse}
calc_rmse_by_horizon <- function(eval_df, by_season = FALSE) {

  df <- eval_df |>
    filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs))

  if (by_season) {
    df <- df |>
      mutate(
        month  = month(reference_date),
        Season = case_when(
          month %in% c(12, 1, 2)  ~ "Winter",
          month %in% c(3,  4, 5)  ~ "Spring",
          month %in% c(6,  7, 8)  ~ "Summer",
          month %in% c(9, 10, 11) ~ "Fall"
        )
      )
    df |>
      group_by(model_id, Horizon, Season) |>
      summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)), .groups = "drop")
  } else {
    df |>
      group_by(model_id, Horizon) |>
      summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)), .groups = "drop")
  }
}
```








## RMSE Calculations

### Overall RMSE by horizon

```{r rmse-overall}
# # Shared label map (1m) — used in RMSE chunks, flow chunks, and example forecast
# model_labels_1m <- c(
#   "M1_1m_lag"          = "AR only",
#   "M2_1m_lag_LAKEb"    = "AR + in-lake",
#   "M4_1m_lag_LAKEb_RH" = "AR + in-lake + catchment",
#   "M5_1m_lag_RH"       = "AR + catchment"
# )
# model_labels <- model_labels_1m          # active map (reset before flow chunks)
# models_keep  <- names(model_labels)
# 
# # Shared color palette and legend factor order — used in every model_id plot
# model_colors <- c(
#   "AR only"                  = "#C77CFF",
#   "AR + in-lake"             = "#00BA38",
#   "AR + catchment"           = "#F8766D",
#   "AR + in-lake + catchment" = "#619CFF"
# )
# model_levels <- names(model_colors)

# rmse_all <- bind_rows(
#   calc_rmse_by_horizon(m1_1m),
#   calc_rmse_by_horizon(m2b_1m),
#   calc_rmse_by_horizon(m4b_1m),
#   calc_rmse_by_horizon(m5_1m)
# ) |>
#   mutate(model_id = factor(recode(model_id, !!!model_labels), levels = model_levels))


rmse_all <- calc_rmse_by_horizon(Pred_eval_df)

rmse_all |> 
  mutate(model_class = sub("_.*", "", model_id)) |> 
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |> 
  ggplot(aes(x = Horizon, y = RMSE, col = model_vars)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_wrap(~model_class)+
  # scale_color_manual(values = model_colors) +
  labs(x = "Prediction horizon (days ahead)", y = "RMSE (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))

```

### Seasonal RMSE by horizon

```{r rmse-seasonal}
rmse_all_season <- bind_rows(
  calc_rmse_by_horizon(m1_1m,  by_season = TRUE),
  calc_rmse_by_horizon(m2b_1m, by_season = TRUE),
  calc_rmse_by_horizon(m4b_1m, by_season = TRUE),
  calc_rmse_by_horizon(m5_1m,  by_season = TRUE)
) |>
  mutate(model_id = factor(recode(model_id, !!!model_labels), levels = model_levels))

ggplot(rmse_all_season, aes(x = horizon, y = RMSE, col = model_id)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~Season, nrow = 1) +
  scale_color_manual(values = model_colors) +
  labs(title = "Prediction skill by horizon and season",
       x = "Prediction horizon (days ahead)", y = "RMSE (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))
```


---

## Model Skill Relative to Persistence

Skill scores express improvement over the "AR only" model (treated as persistence here).
A skill of 0 means no improvement; positive values indicate better-than-persistence skill;
negative values indicate worse performance. Calculated as:

**Skill = 1 − (model RMSE / AR-only RMSE)**

```{r skill-calc}
skill_rmse <- Pred_eval_df |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |> 
  mutate(month  = month(forecast_date),
        Season = case_when(
          month %in% c(12, 1, 2)  ~ "Winter",
          month %in% c(3,  4, 5)  ~ "Spring",
          month %in% c(6,  7, 8)  ~ "Summer",
          month %in% c(9, 10, 11) ~ "Fall")) |> 
  group_by(model_id, Horizon) |>
  # group_by(model_id, Horizon, Season) |>
  summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)), .groups = "drop") |>
  pivot_wider(names_from = model_id, values_from = RMSE) |>
  # select(horizon, `AR + in-lake`, `AR + catchment`, `AR + in-lake + catchment`) |>
  # pivot_longer(-horizon, names_to = "model_id", values_to = "skill") |> 
    select(Season, horizon, `AR + in-lake`, `AR + catchment`, `AR + in-lake + catchment`) |>
  pivot_longer(-c(Season, horizon), names_to = "model_id", values_to = "skill") |>
  mutate(model_id = factor(model_id, levels = model_levels))



  
#NEW
rmse_by_horizon <- Pred_eval_df |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |>
  group_by(model_id, Horizon) |>
  summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
            .groups = "drop")

persistence_rmse <- rmse_by_horizon |>
  filter(model_id == "Persistence") |>
  select(Horizon, RMSE_persistence = RMSE)

skill_by_horizon <- rmse_by_horizon |>
  filter(model_id != "Persistence") |>
  left_join(persistence_rmse, by = "Horizon") |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

#PLOT
skill_by_horizon |> 
  mutate(model_class = sub("_.*", "", model_id)) |> 
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |> 
ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~model_class)+
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(x = "Forecast Horizon (days)", y = "Skill (vs. Persistence)",
       color = NULL) +
  theme_bw() +
  theme(legend.position = "top")

```
 
```{r skill-plot}
ggplot(skill_rmse, aes(x = horizon, y = skill, col = model_id)) +
  geom_hline(yintercept = 0, linewidth = 1, linetype = "dashed", color = "black") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  facet_wrap(~factor(Season, levels = c("Spring", "Summer", "Fall", "Winter")), nrow = 1)+
  scale_color_manual(values = model_colors) +
  labs(
    #title = "Forecast skill relative to AR-only (persistence)",
    x     = "Prediction horizon (days ahead)",
    y     = "Skill score",
    col   = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 24))

```


