---
title: "Evaluate Predictions"
author: "DWH"
date: "2026-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages and Data

```{r packages}
library(tidyverse)
library(lubridate)
library(Metrics) #rmse and mae
library(patchwork)
```


```{r load-data}
#bring in observed data for eval
eval <- read_csv("../Data/Daily_catwalk_RH_2021_2026.csv") |>
  mutate(Date = as.Date(Date)) |> select(Date, fDOM_1_QSU_daily) |> 
  rename(fDOM_1m_obs = fDOM_1_QSU_daily)


##4 cast outputs 
AR <- read_csv("../Data/AR_forecasts_2024-2025_27may26.csv")

ARIMA <- read_csv("../Data/ARIMA_forecasts_2024-2025_27may26.csv")

XGB <- read_csv("../Data/XGB_forecasts_2024-2025_27may26.csv")

NNETAR <- read_csv("../Data/NNETAR_forecasts_2024-2025_27may26.csv")

Persistence <- read_csv("../Data/Persistence_forecasts_2024-2025_27may26.csv")


#join predictions 
predictions <- rbind(AR, ARIMA, Persistence, XGB, NNETAR) 

unique(predictions$model_id)


#ensemble
predict_ensemble <- predictions |> 
 mutate(model_class = sub("_.*", "", model_id),
         model_vars  = sub("^[^_]*_[^_]*_", "", model_id))  |> 
  filter(model_id != "Persistence") |> 
  group_by(reference_date, forecast_date, Horizon, model_vars) |> 
  summarise(fDOM_1m_predicted = mean(fDOM_1m_predicted, na.rm = T),
            .groups = "drop") |> 
  mutate(model_id = paste0("Ensemble_lag_", model_vars)) |> 
  select(-model_vars)


predictions <- rbind(predictions, predict_ensemble) |> 
   mutate(model_vars  = sub("^[^_]*_[^_]*_", "", model_id))  |> 
    filter(model_vars != "only") |> 
    select(-model_vars)


unique(predictions$model_id)

#### join
Pred_eval_df <- left_join(predictions, eval, by = c("forecast_date" =  "Date" )) |> 
  mutate(model_class = sub("_.*", "", model_id)) |> 
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |> 
  select(model_id, model_class, model_vars,
         reference_date, forecast_date, Horizon,
         fDOM_1m_predicted, fDOM_1m_obs
         )


##set model colors 
model_vars_colors <- c("lake" = "#56B4E9", "catchment" = "#009E73", "lake_catchment" = "#CC79A7", "Persistence" = "black")

#c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")


```



## RMSE and Skill across full TS

RMSE and Skill for ensemble: MS Figure 3

```{r}

## calc eval metrics across all models 
metrics_eval_all <- Pred_eval_df |> 
  #remove NAs and calc RMSE
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |> 
  group_by(model_id, model_class, model_vars, Horizon) |> 
  summarise(RMSE = Metrics::rmse(fDOM_1m_obs, fDOM_1m_predicted),
            MAE = Metrics::mae(fDOM_1m_obs, fDOM_1m_predicted),
            r2 = cor(fDOM_1m_obs, fDOM_1m_predicted)^2,
            .groups = "drop"
            ) 

#pull out persistence to join for skill
rmse_persistence <- metrics_eval_all |> 
  filter(model_class == "Persistence") |> 
  select(Horizon, RMSE_persistence = RMSE)

#### RMSE across full time
rmse_fig3 <- metrics_eval_all |> 
  filter(model_class == "Ensemble") |> 
  # filter(model_class %in% c("Ensemble", "Persistence")) |> 
  ggplot(aes(x = Horizon, y = RMSE, col = model_vars)) +
  geom_line(linewidth = 1) +
  # facet_wrap(~model_class)+
  scale_color_manual(values = model_vars_colors) +
  ylim(0.5, 3.2)+
  labs(x = "Prediction horizon (days)", y = "RMSE (QSU)", col = NULL) +
  annotate("text", x = Inf, y = Inf,  label = "Less accurate", 
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  annotate("text", x = Inf, y = -Inf, label = "More accurate", 
           hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))

rmse_fig3

#### Skill across full time
skill_fig3 <- metrics_eval_all |> 
  filter(model_class == "Ensemble") |> 
  left_join(rmse_persistence, by = "Horizon") |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence)) |> 
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1) +
  ylim(-0.05, 0.25)+
  # facet_wrap(~model_class)+
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 1.5) +
  scale_color_manual(values = model_vars_colors)+
  labs(x = "Prediction Horizon (days)", y = "Skill", color = NULL) +
  annotate("text", x = Inf, y = Inf,  label = "Better than null", 
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  annotate("text", x = Inf, y = -Inf, label = "Worse than null", 
           hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 14))

skill_fig3


##join figure 3

rmse_fig3 | skill_fig3

# ggpubr::ggarrange(rmse_fig3, skill_fig3, ncol = 2, labels = "auto")

```


RMSE and Skill by model: MS Figure 3

```{r}

#### RMSE by model
metrics_eval_all |> 
  # filter(model_class == "Ensemble") |> 
  # filter(model_class %in% c("Ensemble", "Persistence")) |> 
  ggplot(aes(x = Horizon, y = RMSE, col = model_vars)) +
  geom_line(linewidth = 1) +
  facet_wrap(~model_class)+
  scale_color_manual(values = model_vars_colors) +
  # ylim(0.5, 3.2)+
  labs(x = "Prediction horizon (days)", y = "RMSE (QSU)", col = NULL) +
  # annotate("text", x = Inf, y = Inf,  label = "Less accurate", 
  #          hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  # annotate("text", x = Inf, y = -Inf, label = "More accurate", 
  #          hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))



#### skill by model
metrics_eval_all |> 
  filter(model_class != "Persistence") |> 
  left_join(rmse_persistence, by = "Horizon") |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence)) |> 
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1) +
  # ylim(-0.05, 0.25)+
  facet_wrap(~model_class)+
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 1.5) +
  scale_color_manual(values = model_vars_colors)+
  labs(x = "Prediction Horizon (days)", y = "Skill", color = NULL) +
  # annotate("text", x = Inf, y = Inf,  label = "Better than null", 
  #          hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  # annotate("text", x = Inf, y = -Inf, label = "Worse than null", 
  #          hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 14))



```




## Stratfication based on Density diff 

```{r}
## get strat classification
strat <- read_csv("../Data/Daily_catwalk_RH_2021_2026.csv") |> 
  select(Date, Diff_Dens_1_max) |> 
  mutate(Strat = ifelse(Diff_Dens_1_max > 0.1, "Strat", "Mixed"))


## calc eval metrics across all models 
metrics_eval_STRAT <- Pred_eval_df |> 
  left_join(strat, by = c("forecast_date" = "Date")) |> 
  #remove NAs and calc RMSE
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
         !is.na(Strat)) |> 
  group_by(model_id, model_class, model_vars, Horizon, Strat) |> 
  summarise(RMSE = Metrics::rmse(fDOM_1m_obs, fDOM_1m_predicted),
            MAE = Metrics::mae(fDOM_1m_obs, fDOM_1m_predicted),
            r2 = cor(fDOM_1m_obs, fDOM_1m_predicted)^2,
            .groups = "drop"
            ) 

## pull out persistence to join for skill
rmse_persistence_STRAT <- metrics_eval_STRAT |> 
  filter(model_class == "Persistence") |> 
  select(Horizon, Strat, RMSE_persistence = RMSE)

#### RMSE across full time
rmse_STRAT <- metrics_eval_STRAT |> 
  filter(model_class %in% c("Ensemble")) |> 
  ggplot(aes(x = Horizon, y = RMSE, col = model_vars)) +
  geom_line(linewidth = 1) +
  facet_wrap(~Strat)+
  scale_color_manual(values = model_vars_colors) +
  # ylim(0.5, 3.2)+
  labs(x = "Prediction horizon (days)", y = "RMSE (QSU)", col = NULL) +
  annotate("text", x = Inf, y = Inf,  label = "Less accurate", 
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  annotate("text", x = Inf, y = -Inf, label = "More accurate", 
           hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))

rmse_STRAT

#### Skill across full time
skill_STRAT <- metrics_eval_STRAT |> 
  filter(model_class %in% c("Ensemble")) |> 
  left_join(rmse_persistence_STRAT, by = c("Horizon", "Strat")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence)) |> 
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1) +
  # ylim(-0.05, 0.25)+
  facet_wrap(~Strat)+
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 1.5) +
  scale_color_manual(values = model_vars_colors)+
  labs(x = "Prediction Horizon (days)", y = "Skill", color = NULL) +
  annotate("text", x = Inf, y = Inf,  label = "Better than null", 
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  annotate("text", x = Inf, y = -Inf, label = "Worse than null", 
           hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 14))

skill_STRAT

## strats joined

rmse_STRAT | skill_STRAT



```



## High v low Flow 
based on USGS tinker gauge

get tinker data and join

```{r}
##USGS data
USGS_daily <- dataRetrieval::readNWISdv(
  siteNumbers = "02055100",
  parameterCd = "00060",          # discharge in cfs
  startDate = "2023-01-01",
  endDate   = "2026-04-01"
) |>  dataRetrieval::renameNWISColumns()

#set up flow percentile
flow_percentiles <- USGS_daily |>
  filter(Date >= ymd("2024-01-01"), Date <= ymd("2026-01-31")) |>
  filter(!is.na(Flow)) |>
  mutate(flow_percentile = percent_rank(Flow) * 100)

# flag top and bottom 10th percentile dates
flow_flags <- flow_percentiles |>
  mutate(
    decile     = ntile(Flow, 10),
    flow_class = case_when(
      decile == 1  ~ "Low flow",
      decile == 10 ~ "High flow",
      TRUE         ~ "Normal"
    )
  ) |>
  select(Date, Flow, flow_percentile, decile, flow_class)

flow_flag_check <- flow_flags |> 
  filter(flow_class != "Normal")

## Join to pred eval and plot
DryWet2 <- Pred_eval_df |>
  left_join(flow_flags, by = c("forecast_date" = "Date")) |> 
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
          flow_class != "Normal") |>
  group_by(model_id, model_class, model_vars, flow_class, Horizon) |>
  summarise(RMSE = Metrics::rmse(fDOM_1m_obs, fDOM_1m_predicted),
            MAE = Metrics::mae(fDOM_1m_obs, fDOM_1m_predicted),
            r2 = cor(fDOM_1m_obs, fDOM_1m_predicted)^2,
            .groups = "drop"
            ) 

```


RMSE and skill

```{r}

#RMSE from ensemble
DryWet2_ens <- DryWet2 |> 
  filter(grepl("Ensemble", model_id)) 
  
flowrmse2 <- DryWet2_ens |>
  ggplot(aes(x = Horizon, y = RMSE, color = model_vars)) +
  geom_line(linewidth = 1) +
  facet_wrap(~flow_class) +
  labs(x = "Prediction Horizon (days)", y = "RMSE (QSU)", color = NULL) +
    annotate("text", x = Inf, y = Inf,  label = "Less accurate", 
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  annotate("text", x = Inf, y = -Inf, label = "More accurate", 
           hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  scale_color_manual(values = model_vars_colors)+
  theme_bw() +  theme(legend.position = "top", text = element_text(size = 14))


flowrmse2

#### SKill
DryWet2_per <- DryWet2 |> 
  filter(grepl("Persistence", model_id)) |> 
  select(Horizon, flow_class, RMSE_persistence = RMSE)

skill_DryWet2 <- DryWet2_ens |>
  left_join(DryWet2_per, by = c("flow_class", "Horizon")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

#plot
flowskill2 <- skill_DryWet2 |>
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~flow_class) +
  # ylim(-3.5, 1)+
  labs(x = "Prediction Horizon (days)", y = "Skill", color = NULL) +
    annotate("text", x = Inf, y = Inf,  label = "Better than null", 
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  annotate("text", x = Inf, y = -Inf, label = "Worse than null", 
           hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  scale_color_manual(values = model_vars_colors)+
  theme_bw() +  theme(legend.position = "top", text = element_text(size = 14))


flowskill2


#stack plots

flowrmse2 / flowskill2

```


## High Variability days

Set up data for high var days 

```{r}
#Firs first difference value > 90th percentile
eval2 <- eval |> mutate(fDOM_1dif = abs(fDOM_1m_obs - lag(fDOM_1m_obs, 1)))

summary(eval2$fDOM_1dif)
quantile(eval2$fDOM_1dif, probs = 0.9, na.rm = TRUE)


eval2$HighVar = ifelse(eval2$fDOM_1dif > 0.68, "High_Var", "Low_Var" )

highvar_df <- eval2 |> select(Date, HighVar)


```


```{r}
#### RMSE
rmse_HighVar <-  Pred_eval_df |> 
  left_join(highvar_df, by = c("forecast_date" = "Date")) |> 
  filter(grepl("Ensemble", model_id))  |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
         !is.na(HighVar)) |>
  group_by(model_id, model_vars, model_class, HighVar, Horizon) |>
  summarise(RMSE = Metrics::rmse(fDOM_1m_obs, fDOM_1m_predicted),
            .groups = "drop"
            ) 

highvar_rmse <- rmse_HighVar |>
  ggplot(aes(x = Horizon, y = RMSE, color = model_vars)) +
  geom_line(linewidth = 1) +
  facet_wrap(~HighVar) +
  labs(x = "Prediction Horizon (days)", y = "RMSE (QSU)", color = NULL) +
    annotate("text", x = Inf, y = Inf,  label = "Less accurate", 
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  annotate("text", x = Inf, y = -Inf, label = "More accurate", 
           hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  scale_color_manual(values = model_vars_colors)+
  theme_bw() +  theme(legend.position = "top", text = element_text(size = 14))

highvar_rmse


#### SKill
persistence_HighVar <- Pred_eval_df |>
  left_join(highvar_df, by = c("forecast_date" = "Date")) |>
  filter(model_id == "Persistence") |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
          !is.na(HighVar)) |>
  group_by(HighVar, Horizon) |>
  summarise(RMSE_persistence = Metrics::rmse(fDOM_1m_obs, fDOM_1m_predicted),
            .groups = "drop")

skill_HighVar <- rmse_HighVar |>
  left_join(persistence_HighVar, by = c("HighVar", "Horizon")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

highvar_skill <- skill_HighVar |>
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~HighVar) +
  labs(x = "Prediction Horizon (days)", y = "Skill", color = NULL) +
    annotate("text", x = Inf, y = Inf,  label = "Better than null", 
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic", color = "grey40") +
  annotate("text", x = Inf, y = -Inf, label = "Worse than null", 
           hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic", color = "grey40") +
  scale_color_manual(values = model_vars_colors)+
  theme_bw() +  theme(legend.position = "top", text = element_text(size = 14))

highvar_skill


##plot both

highvar_rmse / highvar_skill

```


## Skill grouping MS figure 

```{r}

skill_STRAT / flowskill2 / highvar_skill

```


## RMSE grouping SI figure 

```{r}

rmse_STRAT / flowrmse2 / highvar_rmse

```


################### OLD ############################################



### Forecast visualisation: `plot_forecast()`

Plots observed history leading up to `forecast_date`, model prediction lines + points
for each requested model, and (optionally) observed values during the forecast window.

**Arguments:**
- `forecast_date` — the reference date the forecast was issued from
- `models` — character vector of `model_id` values to include
- `Pred_eval_df` — the joined prediction + observation dataframe
- `obs_df` — observed timeseries (`eval` in this script; needs `Date` + `fDOM_1m_obs`)
- `history_days` — how many days of pre-forecast observed history to show (default 60)
- `show_horizon_obs` — if `TRUE` (default), overlays observed fDOM during the forecast
  window as open black circles; set to `FALSE` to hide them

```{r fn-plot-forecast}
plot_forecast <- function(forecast_date,
                          models,
                          Pred_eval_df,
                          obs_df,
                          history_days     = 3,
                          show_horizon_obs = TRUE) {

  ref_date <- as.Date(forecast_date)

  # Observed history: history_days before forecast_date through forecast_date
  history_obs <- obs_df |>
    filter(Date >= ref_date - history_days,
           Date <= ref_date)

  # Model predictions for this reference date and selected models
  fc_preds <- Pred_eval_df |>
    filter(reference_date == ref_date,
           model_id %in% models) |>
    select(model_id, fc_date = forecast_date, Horizon, fDOM_1m_predicted, fDOM_1m_obs)

  if (nrow(fc_preds) == 0) {
    warning("No forecast rows found for reference_date = ", ref_date,
            " and the requested models.")
    return(NULL)
  }

  # Horizon observations — unique across models, for optional overlay
  horizon_obs <- fc_preds |>
    distinct(fc_date, fDOM_1m_obs) |>
    filter(!is.na(fDOM_1m_obs))

  p <- ggplot() +
    # Pre-forecast observed history
     geom_point(data = history_obs,
               aes(x = Date, y = fDOM_1m_obs),
               color = "black", size = 1.5) +
    # Vertical dashed line marking the forecast start
    geom_vline(xintercept = ref_date,
               linetype = "dashed", linewidth = 0.8) +
    # Model forecast lines + points
    geom_line(data  = fc_preds,
              aes(x = fc_date, y = fDOM_1m_predicted, color = model_id),
              linewidth = 0.9) +
    geom_point(data = fc_preds,
               aes(x = fc_date, y = fDOM_1m_predicted, color = model_id),
               size = 1.8) +
    labs(title = paste0("30-day forecast issued ", ref_date),
         x = "Date", y = "fDOM (QSU)", color = NULL) +
    theme_bw() +
    theme(legend.position = "top", text = element_text(size = 13))

  # Optional: overlay observed values during the forecast horizon
  if (show_horizon_obs) {
    p <- p +
      geom_point(data  = horizon_obs,
                 aes(x = fc_date, y = fDOM_1m_obs),
                 color = "black", shape = 21, fill = "white",
                 size = 1.5, stroke = 1.2)
  }

  p
}
```

### Example usage

```{r plot-forecast-example}

models_ensemble_persist <- c("Ensemble_lag_lake", "Ensemble_lag_catchment", 
                             "Ensemble_lag_lake_catchment", "Persistence")

models_AR_persist <- c("AR_lag_lake", "AR_lag_catchment", 
                             "AR_lag_lake_catchment", "Persistence")


# All models for a single date
plot_forecast(
  forecast_date    = "2024-06-01",
  models           = unique(Pred_eval_df$model_id),
  Pred_eval_df     = Pred_eval_df,
  obs_df           = eval,
  history_days     = 3,
  show_horizon_obs = TRUE
)

# Subset of models, no horizon observations
plot_forecast(
  forecast_date    = "2025-02-01",
  models           = models_ensemble_persist,
  Pred_eval_df     = Pred_eval_df,
  obs_df           = eval,
  history_days     = 3,
  show_horizon_obs = T
)


## Interesting dates: "2025-09-15";

```


---

## PDF: One plot per reference date

```{r fn-save-forecast-pdf}
save_forecast_pdf <- function(Pred_eval_df,
                               models,
                               output_path  = "../Output/forecast_all_dates.pdf",
                               plots_per_page = 12,
                               ncol           = 3,
                               width          = 16, height = 20) {

  ref_dates <- Pred_eval_df |>
    filter(model_id %in% models) |>
    pull(reference_date) |>
    unique() |>
    sort()

  # Build one ggplot per reference date
  plot_list <- lapply(ref_dates, function(d) {
    d <- as.Date(d)

    plot_df <- Pred_eval_df |>
      filter(reference_date == d, model_id %in% models)

    obs_df <- plot_df |>
      distinct(forecast_date, fDOM_1m_obs) |>
      filter(!is.na(fDOM_1m_obs))

    ggplot() +
      geom_line(data  = plot_df,
                aes(x = forecast_date, y = fDOM_1m_predicted, color = model_id),
                linewidth = 0.7) +
      geom_point(data = plot_df,
                 aes(x = forecast_date, y = fDOM_1m_predicted, color = model_id),
                 size = 1.2) +
      geom_point(data  = obs_df,
                 aes(x = forecast_date, y = fDOM_1m_obs),
                 color = "black", size = 1.8) +
      ylim(1,22) +
      labs(title = format(d, "%b %d %Y"), x = NULL, y = "fDOM (QSU)", color = NULL) +
      theme_bw() +
      theme(legend.position  = "top",
            legend.text      = element_text(size = 6),
            legend.key.size  = unit(0.4, "cm"),
            axis.text.x      = element_text(angle = 30, hjust = 1, size = 7),
            axis.text.y      = element_text(size = 7),
            plot.title       = element_text(size = 8),
            axis.title.y     = element_text(size = 8))
  })

  # Split into pages of plots_per_page and print each page
  pages <- split(plot_list, ceiling(seq_along(plot_list) / plots_per_page))

  pdf(output_path, width = width, height = height)

  for (pg in pages) {
    print(wrap_plots(pg, ncol = ncol, guides = "collect") &
            theme(legend.position = "top"))
  }

  dev.off()
  cat("PDF saved to:", output_path, "\n")
}
```

```{r save-pdf}

models_ensemble_persist <- c("Ensemble_lag_lake", "Ensemble_lag_catchment", 
                             "Ensemble_lag_lake_catchment", "Persistence")

save_forecast_pdf(
  Pred_eval_df = Pred_eval_df,
  models       = models_ensemble_persist, #unique(Pred_eval_df$model_id),
  output_path  = "../Data/forecast_all_dates3.pdf"
)
```

















