---
title: "Evaluate Predictions"
author: "DWH"
date: "2026-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages and Data

```{r packages}
library(tidyverse)
library(lubridate)
library(patchwork)
```


```{r load-data}
eval <- read_csv("../Data/Daily_catwalk_RH_2021_2026.csv") |>
  mutate(Date = as.Date(Date)) |> select(Date, fDOM_1_QSU_daily) |> 
  rename(fDOM_1m_obs = fDOM_1_QSU_daily)


##4 cast outputs 
AR <- read_csv("../Data/AR_forecasts_2024-2025_27may26.csv")

ARIMA <- read_csv("../Data/ARIMA_forecasts_2024-2025_27may26.csv")

XGB <- read_csv("../Data/XGB_forecasts_2024-2025_27may26.csv")

NNETAR <- read_csv("../Data/NNETAR_forecasts_2024-2025_27may26.csv")

Persistence <- read_csv("../Data/Persistence_forecasts_2024-2025_27may26.csv")

# RF <- read_csv("../Data/RF_forecasts_2024_22may26.csv")

# Prophet <- read_csv("../Data/Prophet_forecasts_2024_25may26.csv")



#join predictions 
predictions <- rbind(AR, ARIMA, Persistence, XGB, NNETAR) #, XGB, NNETAR
# predictions <- rbind(AR, Persistence)


#ensemble
unique(predictions$model_id)

predict_ensemble <- predictions |> 
 mutate(model_class = sub("_.*", "", model_id),
         model_vars  = sub("^[^_]*_[^_]*_", "", model_id))  |> 
  filter(model_id != "Persistence") |> 
  group_by(reference_date, forecast_date, Horizon, model_vars) |> 
    summarise(fDOM_1m_predicted = median(fDOM_1m_predicted, na.rm = T), ## CHECK MEAN OR MEDIAN; was mean originally
            .groups = "drop") |> 
  # summarise(fDOM_1m_predicted_mean = mean(fDOM_1m_predicted, na.rm = T), ## CHECK MEAN OR MEDIAN; was mean already
  #           fDOM_1m_predicted_median = median(fDOM_1m_predicted, na.rm = T), ## CHECK MEAN OR MEDIAN; was mean already
  #           .groups = "drop") |> 
  # mutate(pred_resid = fDOM_1m_predicted_mean - fDOM_1m_predicted_median) |> 
  mutate(model_id = paste0("Ensemble_lag_", model_vars)) |> 
  select(-model_vars)


predictions <- rbind(predictions, predict_ensemble) |> 
   mutate(model_vars  = sub("^[^_]*_[^_]*_", "", model_id))  |> 
    filter(model_vars != "only") |> 
    select(-model_vars)

unique(predictions$model_id)

#### join
Pred_eval_df <- left_join(predictions, eval, by = c("forecast_date" =  "Date" )) 


##set model colors 
model_vars_colors <- c("lake" = "#56B4E9", "catchment" = "#009E73", "lake_catchment" = "#CC79A7", "Persistence" = "black")

#c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000")


```


## Residual boxplots

```{r}

zzz <- Pred_eval_df |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |>
  mutate(
    model_class = sub("_.*", "", model_id),
    model_vars  = sub("^[^_]*_[^_]*_", "", model_id),
    residual    = abs(fDOM_1m_obs - fDOM_1m_predicted)
  ) |> 
  filter(model_vars == "lake_catchment",
         Horizon > 2, Horizon < 18,
         residual > 6,
         month(forecast_date) %in% c(3,4,5)
         )

zz <- Pred_eval_df |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |>
  mutate(
    model_class = sub("_.*", "", model_id),
    model_vars  = sub("^[^_]*_[^_]*_", "", model_id),
    residual    = abs(fDOM_1m_obs - fDOM_1m_predicted)
  ) |>
    filter(model_class == "Ensemble") |> 
    # filter(!(year(forecast_date) == 2025 & month(forecast_date) == 5)) |>  # remove May 2025
   mutate(month = month(forecast_date),
         Season = case_when(
          month %in% c(12, 1, 2)  ~ "Winter",
          month %in% c(3,  4, 5)  ~ "Spring",
          month %in% c(6,  7, 8)  ~ "Summer",
          month %in% c(9, 10, 11) ~ "Fall"
        )) |> 
   filter(Season %in% c("Spring", "Summer")) |> 
  ggplot(aes(x = factor(Horizon), y = residual, color = model_vars)) +
  geom_boxplot(aes(fill = model_vars), alpha = 0.3, outlier.shape = NA,
               position = position_dodge(width = 0.8)) +
  geom_jitter(aes(group = model_vars), alpha = 0.15, size = 0.6,
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)) +
  #facet_wrap(~model_vars) +
  facet_grid(Season~model_vars) +
  scale_color_manual(values = model_vars_colors) +
  scale_fill_manual(values  = model_vars_colors) +
  labs(x = "Forecast Horizon (days)", y = "Absolute Residual (QSU)",
       color = NULL, fill = NULL) +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 13),
        axis.text.x = element_text(size = 7))

zz
# plotly::ggplotly(zz)


## Look at just mean or median like pltos below but for residual 
Pred_eval_df |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |>
  mutate(
    model_class = sub("_.*", "", model_id),
    model_vars  = sub("^[^_]*_[^_]*_", "", model_id),
    residual    = abs(fDOM_1m_obs - fDOM_1m_predicted)
  ) |> 
  mutate(month  = month(forecast_date),
        Season = case_when(
          month %in% c(12, 1, 2)  ~ "Winter",
          month %in% c(3,  4, 5)  ~ "Spring",
          month %in% c(6,  7, 8)  ~ "Summer",
          month %in% c(9, 10, 11) ~ "Fall"
        )) |> 
  group_by(model_id, model_class, model_vars, Horizon, Season) |> 
  summarise(Resid = mean(abs(fDOM_1m_predicted - fDOM_1m_obs))) |> ## CHECK mean or median
  filter(model_class == "Ensemble") |> 
  ggplot(aes(x = Horizon, y = Resid, col = model_vars)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_grid(model_class~Season)+
  # scale_color_manual(values = model_colors) +
  labs(x = "Prediction horizon (days ahead)", y = "Resid (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))+
  scale_color_manual(values = model_vars_colors)



obsfDOM <- Pred_eval_df |> 
  ggplot(aes(x = forecast_date, y = fDOM_1m_obs))+
  geom_point()+ geom_line()+
  scale_x_date(date_breaks = "2 month", date_labels = "%b %y")

obsfDOM

```





## RMSE Calculations
 `calc_rmse_by_horizon()`
Returns RMSE at each forecast horizon. Set `by_season = TRUE` to additionally
group by season (Winter / Spring / Summer / Fall).

```{r fn-rmse}
calc_rmse_by_horizon <- function(eval_df, by_season = FALSE) {

  df <- eval_df |>
    filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs))

  if (by_season) {
    df <- df |>
      mutate(
        month  = month(forecast_date),
        Season = case_when(
          month %in% c(12, 1, 2)  ~ "Winter",
          month %in% c(3,  4, 5)  ~ "Spring",
          month %in% c(6,  7, 8)  ~ "Summer",
          month %in% c(9, 10, 11) ~ "Fall"
        )
      )
    df |>
      group_by(model_id, Horizon, Season) |>
      summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
                MAE = mean(abs(fDOM_1m_predicted - fDOM_1m_obs), na.rm = T),
                .groups = "drop")
  } else {
    df |>
      group_by(model_id, Horizon) |>
      summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
                MAE = mean(abs(fDOM_1m_predicted - fDOM_1m_obs), na.rm = T),
                .groups = "drop")
  }
}
```


### Overall RMSE by horizon

```{r rmse-overall}

rmse_all <- calc_rmse_by_horizon(Pred_eval_df)

rmse_plot <- rmse_all |> 
  mutate(model_class = sub("_.*", "", model_id)) |> 
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |> 
  filter(model_vars != "only") |> 
  filter(model_class == "Ensemble") |>
  ggplot(aes(x = Horizon, y = RMSE, col = model_vars)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_wrap(~model_class)+
  # scale_color_manual(values = model_colors) +
  labs(x = "Prediction horizon (days ahead)", y = "RMSE (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))+
  scale_color_manual(values = model_vars_colors)

rmse_plot


##with persistence with ensemble
rmse_all |> 
  mutate(model_class = sub("_.*", "", model_id)) |> 
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |> 
  filter(model_vars != "only") |> 
  # filter(model_class == "Ensemble") |> 
  filter(model_class %in% c("Ensemble", "Persistence")) |> 
  ggplot(aes(x = Horizon, y = RMSE, col = model_vars)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  # facet_wrap(~model_class)+
  # scale_color_manual(values = model_colors) +
  labs(x = "Prediction horizon (days ahead)", y = "RMSE (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))+
  scale_color_manual(values = model_vars_colors)


```

### Seasonal RMSE by horizon

```{r rmse-seasonal}

rmse_all_season  <- calc_rmse_by_horizon(Pred_eval_df,  by_season = TRUE)

 rmse_season_plot <- Pred_eval_df |> 
  #filter(!(year(forecast_date) == 2025 & month(forecast_date) == 5)) |>  # remove May 2025
  #filter(!(year(forecast_date) == 2025 & month(forecast_date) == 7)) |>  # remove July 2025
  calc_rmse_by_horizon( by_season = TRUE) |> 
  mutate(model_class = sub("_.*", "", model_id)) |> 
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |> 
  filter(model_vars != "only") |>
  filter(model_class == "Ensemble") |> 
  ggplot(aes(x = Horizon, y = RMSE, col = model_vars)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_grid(model_class~factor(Season, levels = c("Spring", "Summer", "Fall", "Winter")))+
  # scale_color_manual(values = model_colors) +
  labs(x = "Prediction horizon (days ahead)", y = "RMSE (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "right", text = element_text(size = 14))+
  scale_color_manual(values = model_vars_colors)

rmse_season_plot


##ensemble w/ persistence
Pred_eval_df |> 
  calc_rmse_by_horizon( by_season = TRUE) |> 
  mutate(model_class = sub("_.*", "", model_id)) |> 
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |> 
  filter(model_vars != "only") |>
  filter(model_class %in% c("Ensemble", "Persistence")) |> 
  ggplot(aes(x = Horizon, y = RMSE, col = model_vars)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_wrap(~factor(Season, levels = c("Spring", "Summer", "Fall", "Winter")), nrow =1 )+
  labs(x = "Prediction horizon (days ahead)", y = "RMSE (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "right", text = element_text(size = 14))+
  scale_color_manual(values = model_vars_colors)

```


---

## Model Skill Relative to Persistence

Skill scores express improvement over the "AR only" model (treated as persistence here).
A skill of 0 means no improvement; positive values indicate better-than-persistence skill;
negative values indicate worse performance. Calculated as:

**Skill = 1 − (model RMSE / AR-only RMSE)**

```{r skill-calc}
rmse_by_horizon <- Pred_eval_df |>
  #   mutate(year = year(forecast_date)) |> 
  # filter(year == 2025) |> 
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |>
  group_by(model_id, Horizon) |>
  summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
            .groups = "drop")

persistence_rmse <- rmse_by_horizon |>
  filter(model_id == "Persistence") |>
  select(Horizon, RMSE_persistence = RMSE)

skill_by_horizon <- rmse_by_horizon |>
  filter(model_id != "Persistence") |>
  left_join(persistence_rmse, by = "Horizon") |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

#PLOT
skillplot <- skill_by_horizon |> 
  mutate(model_class = sub("_.*", "", model_id)) |> 
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |> 
  filter(model_vars != "only") |> 
  filter(model_class == "Ensemble") |> 
ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~model_class)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Prediction Horizon (days)", y = "Skill (vs. Persistence)",
       color = NULL) +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 14))+
  scale_color_manual(values = model_vars_colors)

skillplot


rmse_plot | skillplot


```
 
 
```{r skill-plot}
rmse_by_horizon_season <- Pred_eval_df |>
  #     mutate(year = year(forecast_date)) |> 
  # filter(year == 2024) |> 
  mutate(month = month(forecast_date),
         Season = case_when(
          month %in% c(12, 1, 2)  ~ "Winter",
          month %in% c(3,  4, 5)  ~ "Spring",
          month %in% c(6,  7, 8)  ~ "Summer",
          month %in% c(9, 10, 11) ~ "Fall"
        )) |> 
  # filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |>
  group_by(model_id, Season, Horizon) |>
  summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2, na.rm = T)),
            .groups = "drop")

persistence_rmse_season <- rmse_by_horizon_season |>
  filter(model_id == "Persistence") |>
  select(Horizon, Season, RMSE_persistence = RMSE)

skill_by_horizon_season <- rmse_by_horizon_season |>
  filter(model_id != "Persistence") |>
  left_join(persistence_rmse_season, by = c("Horizon", "Season")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

#PLOT
skill_season_plot <- skill_by_horizon_season |> 
  mutate(model_class = sub("_.*", "", model_id)) |> 
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |> 
  filter(model_vars != "only") |> 
  filter(model_class == "Ensemble") |> 
ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.2) +
  facet_grid(model_class~factor(Season, levels = c("Spring", "Summer", "Fall", "Winter")))+
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Prediction Horizon (days)", y = "Skill (vs. Persistence)",
       color = NULL) +
  theme_bw() +
  theme(legend.position = "right")+
  scale_color_manual(values = model_vars_colors)

skill_season_plot

rmse_season_plot / skill_season_plot

```


### Skill by year

```{r}
rmse_by_horizon_year <- Pred_eval_df |>
  mutate(Year = year(forecast_date)  ) |> 
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |>
  group_by(model_id, Year, Horizon) |>
  summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
            .groups = "drop")

persistence_rmse_year <- rmse_by_horizon_year |>
  filter(model_id == "Persistence") |>
  select(Horizon, Year, RMSE_persistence = RMSE)

skill_by_horizon_year <- rmse_by_horizon_year |>
  filter(model_id != "Persistence") |>
  left_join(persistence_rmse_year, by = c("Horizon", "Year")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

#PLOT
skill_by_horizon_year |> 
  filter(Year != 2026) |> 
  mutate(model_class = sub("_.*", "", model_id)) |> 
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |> 
  filter(model_vars != "only") |> 
ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.2) +
  facet_grid(model_class~Year)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Prediction Horizon (days)", y = "Skill (vs. Persistence)",
       color = NULL) +
  theme_bw() +
  theme(legend.position = "top")+
  scale_color_manual(values = model_vars_colors)

```



## Dry vs Wet Month Comparison (Aug 2024 vs Feb 2025)

Ensemble model RMSE and skill for the driest month (August 2024) and wettest month
(February 2025).

```{r dry-wet-rmse}
#### RMSE
wet_dry_ens <- Pred_eval_df |>
  filter(grepl("Ensemble", model_id)) |>
  filter((year(forecast_date) == 2024 & month(forecast_date) == 8) |
         (year(forecast_date) == 2025 & month(forecast_date) == 2)) |>
  mutate(Period = case_when(
    year(forecast_date) == 2024 & month(forecast_date) == 8 ~ "Aug 2024 (Dry)",
    year(forecast_date) == 2025 & month(forecast_date) == 2 ~ "Feb 2025 (Wet)"
  ))

rmse_wet_dry <- wet_dry_ens |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |>
  group_by(model_id, Period, Horizon) |>
  summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
             MAE = mean(abs(fDOM_1m_predicted - fDOM_1m_obs), na.rm = T),
            .groups = "drop")

WetDry_rmse <- rmse_wet_dry |>
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |>
  ggplot(aes(x = Horizon, y = RMSE, color = model_vars)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_wrap(~Period) +
  labs(x = "Prediction Horizon (days)", y = "RMSE (QSU)", color = NULL) +
  theme_bw() +
  theme(legend.position = "right", text = element_text(size = 13))+
  scale_color_manual(values = model_vars_colors)


#### SKill
persistence_wet_dry <- Pred_eval_df |>
  filter(model_id == "Persistence") |>
  filter((year(forecast_date) == 2024 & month(forecast_date) == 8) |
         (year(forecast_date) == 2025 & month(forecast_date) == 2)) |>
  mutate(Period = case_when(
    year(forecast_date) == 2024 & month(forecast_date) == 8 ~ "Aug 2024 (Dry)",
    year(forecast_date) == 2025 & month(forecast_date) == 2 ~ "Feb 2025 (Wet)"
  )) |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs)) |>
  group_by(Period, Horizon) |>
  summarise(RMSE_persistence = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
            .groups = "drop")

skill_wet_dry <- rmse_wet_dry |>
  left_join(persistence_wet_dry, by = c("Period", "Horizon")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

WetDry_skill <- skill_wet_dry |>
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |>
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 1.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Period) +
  ylim(-3.5, 1)+
  labs(x = "Prediction Horizon (days)", y = "Skill (vs. Persistence)", color = NULL) +
  theme_bw() +
  theme(legend.position = "right", text = element_text(size = 13))+
  scale_color_manual(values = model_vars_colors)



#stack plots

WetDry_rmse / WetDry_skill

```




## Flow duratin curve

```{r}
flow <- read_csv("../Data/Daily_catwalk_RH_2021_2026.csv")

fdc <- flow |>
  select(Date, RH_Q_m3day) |> 
  filter(!is.na(RH_Q_m3day)) |>
  arrange(desc(RH_Q_m3day)) |>
  mutate(exceedance = row_number() / (n() + 1) * 100)

fdc |> 
  ggplot(aes(x = exceedance, y = RH_Q_m3day)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Exceedance Probability (%)", y = "Q (m3/day)",
       title = "Flow Duration Curve") +
  theme_bw()

```




## Strat v mixed


```{r}

#get strat classification
getwd()
driverdf <- read_csv("../Data/Daily_catwalk_RH_2021_2026.csv")

strat <- driverdf |> 
  select(Date, Diff_Dens_1_max) |> 
  mutate(Strat = ifelse(Diff_Dens_1_max > 0.1, "Strat", "Mixed"))

#### RMSE
Mix_ens <- Pred_eval_df |> 
  left_join(strat, by = c("forecast_date" = "Date")) |> 
  filter(grepl("Ensemble", model_id)) 

rmse_Mix <- Mix_ens |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
         !is.na(Strat)) |>
  group_by(model_id, Strat, Horizon) |>
  summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
            MAE = mean(abs(fDOM_1m_predicted - fDOM_1m_obs), na.rm = T),
            .groups = "drop")

Mix_rmse <- rmse_Mix |>
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |>
  ggplot(aes(x = Horizon, y = RMSE, color = model_vars)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_wrap(~Strat) +
  labs(x = "Prediction Horizon (days)", y = "RMSE (QSU)", color = NULL) +
  theme_bw() +
  theme(legend.position = "right", text = element_text(size = 13))+
  scale_color_manual(values = model_vars_colors)


#### SKill
persistence_Mix <- Pred_eval_df |>
  left_join(strat, by = c("forecast_date" = "Date")) |>
  filter(model_id == "Persistence") |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
          !is.na(Strat)) |>
  group_by(Strat, Horizon) |>
  summarise(RMSE_persistence = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
            .groups = "drop")

skill_Mix <- rmse_Mix |>
  left_join(persistence_Mix, by = c("Strat", "Horizon")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

Mix_skill <- skill_Mix |>
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |>
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 1.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Strat) +
  labs(x = "Prediction Horizon (days)", y = "Skill (vs. Persistence)", color = NULL) +
  theme_bw() +
  theme(legend.position = "right", text = element_text(size = 13))+
  scale_color_manual(values = model_vars_colors)



#stack plots

Mix_rmse / Mix_skill



##resid 
Mix_ens |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
         !is.na(Strat)) |>
  mutate(
    model_class = sub("_.*", "", model_id),
    model_vars  = sub("^[^_]*_[^_]*_", "", model_id),
    residual    = abs(fDOM_1m_obs - fDOM_1m_predicted)
  ) |> 
  filter(model_class == "Ensemble") |> 
  group_by(model_id, model_class, model_vars, Horizon, Strat) |> 
  summarise(Resid = mean(abs(fDOM_1m_predicted - fDOM_1m_obs))) |> ## CHECK mean or median
  ggplot(aes(x = Horizon, y = Resid, col = model_vars)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_grid(model_class~Strat)+
  labs(x = "Prediction horizon (days ahead)", y = "Resid (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))+
  scale_color_manual(values = model_vars_colors)

```



## Density driven seasons 

```{r}

#get strat classification
getwd()
driverdf <- read_csv("../Data/Daily_catwalk_RH_2021_2026.csv")

strat <- driverdf |> 
  select(Date, Diff_Dens_1_max) |> 
  mutate(Strat = ifelse(Diff_Dens_1_max > 0.3, "Strat", NA),
         Strat = ifelse(Diff_Dens_1_max < 0.1, "Mixed", Strat),
         Strat = ifelse(is.na(Strat), "Shoulder", Strat),
          Strat = ifelse(is.na(Diff_Dens_1_max), NA, Strat),
          )


strat |> 
  ggplot(aes(x = Date, y =Strat))+ geom_point()

#### RMSE
Mix_ens <- Pred_eval_df |> 
  left_join(strat, by = c("forecast_date" = "Date")) |> 
  filter(grepl("Ensemble", model_id)) 

rmse_Mix <- Mix_ens |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
         !is.na(Strat)) |>
  group_by(model_id, Strat, Horizon) |>
  summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
            MAE = mean(abs(fDOM_1m_predicted - fDOM_1m_obs), na.rm = T),
            .groups = "drop")

Mix_rmse <- rmse_Mix |>
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |>
  ggplot(aes(x = Horizon, y = RMSE, color = model_vars)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_wrap(~Strat) +
  labs(x = "Prediction Horizon (days)", y = "RMSE (QSU)", color = NULL) +
  theme_bw() +
  theme(legend.position = "right", text = element_text(size = 13))+
  scale_color_manual(values = model_vars_colors)


#### SKill
persistence_Mix <- Pred_eval_df |>
  left_join(strat, by = c("forecast_date" = "Date")) |>
  filter(model_id == "Persistence") |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
          !is.na(Strat)) |>
  group_by(Strat, Horizon) |>
  summarise(RMSE_persistence = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
            .groups = "drop")

skill_Mix <- rmse_Mix |>
  left_join(persistence_Mix, by = c("Strat", "Horizon")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

Mix_skill <- skill_Mix |>
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |>
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 1.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Strat) +
  labs(x = "Prediction Horizon (days)", y = "Skill (vs. Persistence)", color = NULL) +
  theme_bw() +
  theme(legend.position = "right", text = element_text(size = 13))+
  scale_color_manual(values = model_vars_colors)



#stack plots

Mix_rmse / Mix_skill


```



## First difference trial

```{r}
#Firs first difference value > 90th percentile
eval2 <- eval |> mutate(fDOM_1dif = abs(fDOM_1m_obs - lag(fDOM_1m_obs, 1)))

summary(eval2$fDOM_1dif)
quantile(eval2$fDOM_1dif, probs = 0.9, na.rm = TRUE)


eval2$HighVar = ifelse(eval2$fDOM_1dif > 0.68, "High_Var", "Low_Var" )

highvar_df <- eval2 |> select(Date, HighVar)

## Join to predvar and plot

#### RMSE
HighVar_ens <- Pred_eval_df |> 
  left_join(highvar_df, by = c("forecast_date" = "Date")) |> 
  filter(grepl("Ensemble", model_id)) 

rmse_HighVar <- HighVar_ens |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
         !is.na(HighVar)) |>
  group_by(model_id, HighVar, Horizon) |>
  summarise(RMSE = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
            MAE = mean(abs(fDOM_1m_predicted - fDOM_1m_obs), na.rm = T),
            .groups = "drop")

highvar_rmse <- rmse_HighVar |>
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |>
  ggplot(aes(x = Horizon, y = RMSE, color = model_vars)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_wrap(~HighVar) +
  labs(x = "Prediction Horizon (days)", y = "RMSE (QSU)", color = NULL) +
  theme_bw() +
  theme(legend.position = "right", text = element_text(size = 13))+
  scale_color_manual(values = model_vars_colors)


#### SKill
persistence_HighVar <- Pred_eval_df |>
  left_join(highvar_df, by = c("forecast_date" = "Date")) |>
  filter(model_id == "Persistence") |>
  filter(!is.na(fDOM_1m_predicted), !is.na(fDOM_1m_obs),
          !is.na(HighVar)) |>
  group_by(HighVar, Horizon) |>
  summarise(RMSE_persistence = sqrt(mean((fDOM_1m_predicted - fDOM_1m_obs)^2)),
            .groups = "drop")

skill_HighVar <- rmse_HighVar |>
  left_join(persistence_HighVar, by = c("HighVar", "Horizon")) |>
  mutate(Skill = 1 - (RMSE / RMSE_persistence))

highvar_skill <- skill_HighVar |>
  mutate(model_vars = sub("^[^_]*_[^_]*_", "", model_id)) |>
  ggplot(aes(x = Horizon, y = Skill, color = model_vars)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 1.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~HighVar) +
  labs(x = "Prediction Horizon (days)", y = "Skill (vs. Persistence)", color = NULL) +
  theme_bw() +
  theme(legend.position = "right", text = element_text(size = 13))+
  scale_color_manual(values = model_vars_colors)


highvar_rmse / highvar_skill

```







### Forecast visualisation: `plot_forecast()`

Plots observed history leading up to `forecast_date`, model prediction lines + points
for each requested model, and (optionally) observed values during the forecast window.

**Arguments:**
- `forecast_date` — the reference date the forecast was issued from
- `models` — character vector of `model_id` values to include
- `Pred_eval_df` — the joined prediction + observation dataframe
- `obs_df` — observed timeseries (`eval` in this script; needs `Date` + `fDOM_1m_obs`)
- `history_days` — how many days of pre-forecast observed history to show (default 60)
- `show_horizon_obs` — if `TRUE` (default), overlays observed fDOM during the forecast
  window as open black circles; set to `FALSE` to hide them

```{r fn-plot-forecast}
plot_forecast <- function(forecast_date,
                          models,
                          Pred_eval_df,
                          obs_df,
                          history_days     = 3,
                          show_horizon_obs = TRUE) {

  ref_date <- as.Date(forecast_date)

  # Observed history: history_days before forecast_date through forecast_date
  history_obs <- obs_df |>
    filter(Date >= ref_date - history_days,
           Date <= ref_date)

  # Model predictions for this reference date and selected models
  fc_preds <- Pred_eval_df |>
    filter(reference_date == ref_date,
           model_id %in% models) |>
    select(model_id, fc_date = forecast_date, Horizon, fDOM_1m_predicted, fDOM_1m_obs)

  if (nrow(fc_preds) == 0) {
    warning("No forecast rows found for reference_date = ", ref_date,
            " and the requested models.")
    return(NULL)
  }

  # Horizon observations — unique across models, for optional overlay
  horizon_obs <- fc_preds |>
    distinct(fc_date, fDOM_1m_obs) |>
    filter(!is.na(fDOM_1m_obs))

  p <- ggplot() +
    # Pre-forecast observed history
     geom_point(data = history_obs,
               aes(x = Date, y = fDOM_1m_obs),
               color = "black", size = 1.5) +
    # Vertical dashed line marking the forecast start
    geom_vline(xintercept = ref_date,
               linetype = "dashed", linewidth = 0.8) +
    # Model forecast lines + points
    geom_line(data  = fc_preds,
              aes(x = fc_date, y = fDOM_1m_predicted, color = model_id),
              linewidth = 0.9) +
    geom_point(data = fc_preds,
               aes(x = fc_date, y = fDOM_1m_predicted, color = model_id),
               size = 1.8) +
    labs(title = paste0("30-day forecast issued ", ref_date),
         x = "Date", y = "fDOM (QSU)", color = NULL) +
    theme_bw() +
    theme(legend.position = "top", text = element_text(size = 13))

  # Optional: overlay observed values during the forecast horizon
  if (show_horizon_obs) {
    p <- p +
      geom_point(data  = horizon_obs,
                 aes(x = fc_date, y = fDOM_1m_obs),
                 color = "black", shape = 21, fill = "white",
                 size = 1.5, stroke = 1.2)
  }

  p
}
```

### Example usage

```{r plot-forecast-example}

models_ensemble_persist <- c("Ensemble_lag_lake", "Ensemble_lag_catchment", 
                             "Ensemble_lag_lake_catchment", "Persistence")

models_AR_persist <- c("AR_lag_lake", "AR_lag_catchment", 
                             "AR_lag_lake_catchment", "Persistence")


# All models for a single date
plot_forecast(
  forecast_date    = "2024-06-01",
  models           = unique(Pred_eval_df$model_id),
  Pred_eval_df     = Pred_eval_df,
  obs_df           = eval,
  history_days     = 3,
  show_horizon_obs = TRUE
)

# Subset of models, no horizon observations
plot_forecast(
  forecast_date    = "2025-02-01",
  models           = models_ensemble_persist,
  Pred_eval_df     = Pred_eval_df,
  obs_df           = eval,
  history_days     = 3,
  show_horizon_obs = T
)


## Interesting dates: "2025-09-15";

```


---

## PDF: One plot per reference date

```{r fn-save-forecast-pdf}
save_forecast_pdf <- function(Pred_eval_df,
                               models,
                               output_path  = "../Output/forecast_all_dates.pdf",
                               plots_per_page = 12,
                               ncol           = 3,
                               width          = 16, height = 20) {

  ref_dates <- Pred_eval_df |>
    filter(model_id %in% models) |>
    pull(reference_date) |>
    unique() |>
    sort()

  # Build one ggplot per reference date
  plot_list <- lapply(ref_dates, function(d) {
    d <- as.Date(d)

    plot_df <- Pred_eval_df |>
      filter(reference_date == d, model_id %in% models)

    obs_df <- plot_df |>
      distinct(forecast_date, fDOM_1m_obs) |>
      filter(!is.na(fDOM_1m_obs))

    ggplot() +
      geom_line(data  = plot_df,
                aes(x = forecast_date, y = fDOM_1m_predicted, color = model_id),
                linewidth = 0.7) +
      geom_point(data = plot_df,
                 aes(x = forecast_date, y = fDOM_1m_predicted, color = model_id),
                 size = 1.2) +
      geom_point(data  = obs_df,
                 aes(x = forecast_date, y = fDOM_1m_obs),
                 color = "black", size = 1.8) +
      ylim(1,22) +
      labs(title = format(d, "%b %d %Y"), x = NULL, y = "fDOM (QSU)", color = NULL) +
      theme_bw() +
      theme(legend.position  = "top",
            legend.text      = element_text(size = 6),
            legend.key.size  = unit(0.4, "cm"),
            axis.text.x      = element_text(angle = 30, hjust = 1, size = 7),
            axis.text.y      = element_text(size = 7),
            plot.title       = element_text(size = 8),
            axis.title.y     = element_text(size = 8))
  })

  # Split into pages of plots_per_page and print each page
  pages <- split(plot_list, ceiling(seq_along(plot_list) / plots_per_page))

  pdf(output_path, width = width, height = height)

  for (pg in pages) {
    print(wrap_plots(pg, ncol = ncol, guides = "collect") &
            theme(legend.position = "top"))
  }

  dev.off()
  cat("PDF saved to:", output_path, "\n")
}
```

```{r save-pdf}

models_ensemble_persist <- c("Ensemble_lag_lake", "Ensemble_lag_catchment", 
                             "Ensemble_lag_lake_catchment", "Persistence")

save_forecast_pdf(
  Pred_eval_df = Pred_eval_df,
  models       = models_ensemble_persist, #unique(Pred_eval_df$model_id),
  output_path  = "../Data/forecast_all_dates3.pdf"
)
```

















