---
title: "AR fDOM Forecasting — Streamlined Workflow"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5
)
```

---

## Packages and Data

```{r packages}
library(tidyverse)
library(lubridate)
```

```{r load-data}
getwd()
ar_df <- read_csv("../Data/Daily_catwalk_met_RH_2021_2025_lagZS.csv") |>
  mutate(Date = as.Date(Date)) |> 
  select(Date, fDOM_1_QSU_daily, fDOM_1m_lag1, 
         Temp_1_C_daily, DOsat_1_pct_daily, Chla_1_ugL_daily,
         RH_Q_m3day, RH_DOC_mgL)

# Training window: all data before 2024
train_cutoff <- ymd("2024-01-01")

# Prediction window: every day of 2024
pred_dates_2024 <- seq(ymd("2024-01-01"), ymd("2024-12-31"), by = "day")

```

---

## Model Fitting

### Fitting function

Fits a Gaussian GLM on data before `train_cutoff`. Returns the fitted model object,
metadata needed for forecasting, and a training plot of observed vs fitted.

```{r fn-fit}
fit_ar_glm <- function(all_data,
                        train_cutoff,
                        target,
                        lag_col,
                        covariates = NULL,
                        model_name) {

  # Build training data — drop NAs in all model columns
  train_df <- all_data |>
    filter(Date < train_cutoff,
           !is.na(!!sym(target)),
           !is.na(!!sym(lag_col)))

  if (!is.null(covariates)) {
    for (cv in covariates) train_df <- train_df |> filter(!is.na(!!sym(cv)))
  }

  # Build formula and fit GLM
  rhs     <- paste(c(lag_col, covariates), collapse = " + ")
  formula <- as.formula(paste(target, "~", rhs))
  mod     <- glm(formula, data = train_df, family = gaussian, na.action = na.fail)

  # In-sample fitted values
  train_df <- train_df |> mutate(fitted = predict(mod, newdata = train_df))

  # Training plot: observed vs fitted timeseries
  p <- ggplot(train_df) +
    geom_point(aes(x = Date, y = !!sym(target), color = "Observed"),
               size = 1.2, alpha = 0.8) +
    geom_point(aes(x = Date, y = fitted, color = "Fitted"),
               size = 1.2, alpha = 0.8) +
    scale_color_manual(values = c("Observed" = "black", "Fitted" = "dodgerblue"),
                       breaks = c("Observed", "Fitted")) +
    labs(title = paste0(model_name, " — training fit (pre-", year(train_cutoff), ")"),
         x = "Date", y = "fDOM (QSU)", color = NULL) +
    theme_bw() +
    theme(legend.position = "top")

  print(summary(mod))

  list(
    model      = mod,
    plot       = p,
    model_name = model_name,
    target     = target,
    lag_col    = lag_col,
    covariates = covariates
  )
}
```

### Fit 4 models

All models trained on data before `r train_cutoff`.

```{r fit-models, results='hide'}
# M1: lag only
AR_lagonly <- fit_ar_glm(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  lag_col      = "fDOM_1m_lag1",
  covariates   = NULL,
  model_name   = "AR_lag_only"
)

# M2: lag + in-lake (Temp, DO, Chla)
AR_lake <- fit_ar_glm(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  lag_col      = "fDOM_1m_lag1",
  covariates   = c("Temp_1_C_daily", "DOsat_1_pct_daily", "Chla_1_ugL_daily"),
  model_name   = "AR_lag_lake"
)

# M3: lag + catchment (RH discharge and DOC)
AR_catch <- fit_ar_glm(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  lag_col      = "fDOM_1m_lag1",
  covariates   = c("RH_Q_m3day", "RH_DOC_mgL"),
  model_name   = "AR_lag_catchment"
)

# M4: lag + in-lake + catchment
AR_lake_catch <- fit_ar_glm(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  lag_col      = "fDOM_1m_lag1",
  covariates   = c("Temp_1_C_daily", "DOsat_1_pct_daily", "Chla_1_ugL_daily",
                   "RH_Q_m3day", "RH_DOC_mgL"),
  model_name   = "AR_lag_lake_catchment"
)


```

### Training plots

```{r plot-training, fig.height=4}
AR_lagonly$plot
AR_lake$plot
AR_catch$plot
AR_lake_catch$plot
```

AIC
Thing to check: that training data is same for all models (incase of weird NA missing a few for one co-var)
```{r}

AIC(m1_1m$model, m2b_1m$model, m4b_1m$model,  m5_1m$model)

```


---

## 30-Day Forecast Function

Generates a 30-day ahead forecast from a single reference date. Follows the
pre-allocated dataframe + for-loop pattern (dwh77/fDOM_forecasting): the dataframe
is seeded with the observed fDOM value on the reference date, then each row is filled
sequentially with the lag pulled from the prior row's prediction.

Observed covariate values are used at each forecast step (perfect covariate assumption).

```{r fn-forecast}
# #inputs to test function
# forecast_date <- ymd("2024-01-01")
# model_obj <- AR_lake_catch
# all_data <- ar_df
# target  = "fDOM_1_QSU_daily"
# horizon = 30


forecast_fdom_30day <- function(forecast_date,
                                 model_obj,
                                 all_data,
                                 target  = "fDOM_1_QSU_daily",
                                 horizon = 30) {

  lag_col    <- model_obj$lag_col
  covariates <- model_obj$covariates
  model      <- model_obj$model

  # Seed: observed fDOM on the reference date
  curr_fdom <- all_data |>
    filter(Date == forecast_date) |>
    pull(!!sym(target))

  if (length(curr_fdom) == 0 || is.na(curr_fdom)) return(NULL)
  
  # Mask future fDOM — prevents any observed or lagged fDOM on/after forecast_date from being accessible during the forecast loop
  all_data <- all_data |>
    mutate(
      !!sym(target)  := if_else(Date >= forecast_date, NA_real_, !!sym(target)),
      !!sym(lag_col) := if_else(Date >  forecast_date, NA_real_, !!sym(lag_col))
    )

  # Sequence of dates: reference date through 30 days ahead
  forecast_dates <- seq(from = forecast_date,
                        to   = forecast_date + horizon,
                        by   = "day")

  # Pre-allocate forecast dataframe
  forecast_df <- tibble(
    model_id =  model_obj$model_name,
    reference_date    = forecast_date,
    forecast_date     = forecast_dates,
    Horizon           = as.integer(forecast_dates - reference_date),
    fDOM_1m_predicted = as.double(NA)
  )

  # Seed horizon 0 with the observed value (used as lag for step 1)
  forecast_df$fDOM_1m_predicted[1] <- curr_fdom

  # Step forward one day at a time
  for (i in 2:length(forecast_dates)) {

    # Lag: prior step's prediction pulled from the forecast dataframe
    fdom_lag <- forecast_df$fDOM_1m_predicted[i - 1]

    # Covariates: observed values for this forecast date
    fc_row <- all_data |> filter(Date == forecast_dates[i])
    if (nrow(fc_row) == 0) next

    # Assemble newdata for predict()
    newdata <- setNames(data.frame(fdom_lag), lag_col)
    if (!is.null(covariates)) {
      for (cv in covariates) newdata[[cv]] <- fc_row[[cv]]
    }

    forecast_df$fDOM_1m_predicted[i] <- as.numeric(predict(model, newdata = newdata))
  }

  # Return only the 30 forecast days (drop horizon 0 seed row)
  forecast_df |> filter(Horizon > 0)
} #end of function 


# #test function
# zz <- forecast_fdom_30day(forecast_date = ymd("2024-01-01"), model_obj = AR_lake_catch,
#                     all_data = ar_df, target  = "fDOM_1_QSU_daily", horizon = 30)



```

---

## Run Forecasts Across 2024

`run_forecasts()` loops `forecast_fdom_30day()` over every valid prediction date
(days where observed fDOM exists to seed the lag) and appends all outputs together.

```{r fn-loop}
run_forecasts <- function(model_obj,
                           all_data,
                           pred_dates,
                           target  = "fDOM_1_QSU_daily",
                           horizon = 30) {

  # Keep only dates with an observed target value (needed to seed the lag)
  valid_dates <- all_data |>
    filter(Date %in% pred_dates, !is.na(!!sym(target))) |>
    pull(Date)

  message(model_obj$model_name, " | forecasting ", length(valid_dates),
          " reference dates...")

  map_dfr(valid_dates, \(d) forecast_fdom_30day(
    forecast_date = d,
    model_obj     = model_obj,
    all_data      = all_data,
    target        = target,
    horizon       = horizon
  )) 
}
```

```{r run-forecasts}
forecasts_m1 <- run_forecasts(AR_lagonly, ar_df, pred_dates_2024)
forecasts_m2 <- run_forecasts(AR_lake, ar_df, pred_dates_2024)
forecasts_m3 <- run_forecasts(AR_catch, ar_df, pred_dates_2024)
forecasts_m4 <- run_forecasts(AR_lake_catch, ar_df, pred_dates_2024)

# Combine all models into one dataframe
all_AR_forecasts <- bind_rows(forecasts_m1, forecasts_m2, forecasts_m3, forecasts_m4)

glimpse(all_AR_forecasts)

```

---

## Save Output

```{r save}
write_csv(all_AR_forecasts,
          "../Data/AR_forecasts_2024_22may26.csv")

message("Saved ", nrow(all_forecasts), " forecast rows across ",
        n_distinct(all_forecasts$model_id), " models.")
```




## Run persistence

```{r}
forecast_persistence <- function(forecast_date,
                                  all_data,
                                  target  = "fDOM_1_QSU_daily",
                                  horizon = 30) {

  curr_fdom <- all_data |>
    filter(Date == forecast_date) |>
    pull(!!sym(target))

  if (length(curr_fdom) == 0 || is.na(curr_fdom)) return(NULL)

  tibble(
    model_id          = "Persistence",
    reference_date    = forecast_date,
    forecast_date     = forecast_date + 1:horizon,
    Horizon           = 1:horizon,
    fDOM_1m_predicted = curr_fdom
  )
}

# Run across 2024
valid_dates <- ar_df |>
  filter(Date %in% pred_dates_2024, !is.na(fDOM_1_QSU_daily)) |>
  pull(Date)

all_persistence_forecasts <- map_dfr(valid_dates, \(d)
  forecast_persistence(forecast_date = d, all_data = ar_df))


write_csv(all_persistence_forecasts,
          "../Data/Persistence_forecasts_2024_22may26.csv")


```














