---
title: "XGBoost fDOM Forecasting — Streamlined Workflow"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5
)
```

---

## Packages and Data

```{r packages}
library(tidyverse)
library(lubridate)
library(tidymodels)
library(vip)
tidymodels_prefer()
```

```{r load-data}
getwd()
ar_df <- read_csv("../Data/Daily_catwalk_met_RH_2021_2025_lagZS.csv") |>
  mutate(Date = as.Date(Date)) |>
  select(Date, fDOM_1_QSU_daily, fDOM_1m_lag1,
         Temp_1_C_daily, DOsat_1_pct_daily, Chla_1_ugL_daily,
         RH_Q_m3day, RH_DOC_mgL)

# Training window: all data before 2024
train_cutoff <- ymd("2024-01-01")

# Prediction window: every day of 2024
pred_dates_2024 <- seq(ymd("2024-01-01"), ymd("2024-12-31"), by = "day")
```

---

## Model Fitting


non function fit 

```{r}
set.seed(100)

target     <- "fDOM_1_QSU_daily"
lag_col    <- "fDOM_1m_lag1"
covariates <- c("Temp_1_C_daily", "DOsat_1_pct_daily", "Chla_1_ugL_daily",
                "RH_Q_m3day", "RH_DOC_mgL")
feature_cols <- c(lag_col, covariates)

train_df_full <- ar_df |>
  filter(Date < train_cutoff) |>
  select(Date, all_of(c(target, feature_cols))) |>
  na.omit()

train_df <- train_df_full |> select(-Date)

xgb_spec <- boost_tree(trees = 500, tree_depth = 4,
                        learn_rate = 0.05, min_n = 10) |>
  set_engine("xgboost") |>
  set_mode("regression")

xgb_wf <- workflow() |>
  add_recipe(recipe(as.formula(paste(target, "~ .")), data = train_df)) |>
  add_model(xgb_spec)

xgb_fit <- xgb_wf |> fit(data = train_df)

# check what we get from the engine
xgb_engine <- extract_fit_engine(xgb_fit)
class(xgb_engine)

# try importance
xgboost::xgb.importance(model = xgb_engine, feature_names = feature_cols)

```




### Fitting function

Fits an XGBoost model via parsnip (Melofton et al. approach) on data before
`train_cutoff`. Hyperparameters are fixed: `trees = 500`, `tree_depth = 4`,
`learn_rate = 0.05`, `min_n = 10` — tuning can be swapped in later.

The fitted object is a tidymodels `workflow_fit`, which keeps the preprocessing
recipe and model together. `predict(workflow_fit, new_data = ...)` returns a
tibble with a `.pred` column.

Returns the fitted workflow, metadata for forecasting, a training plot of
observed vs fitted, and a variable importance plot.

```{r fn-fit}
fit_xgb_model <- function(all_data,
                            train_cutoff,
                            target,
                            lag_col,
                            covariates = NULL,
                            model_name,
                            trees      = 500,
                            tree_depth = 4,
                            learn_rate = 0.05,
                            min_n      = 10,
                            seed       = 100) {

  set.seed(seed)
  feature_cols <- c(lag_col, covariates)

  # Build training data — keep Date for plotting, drop it before fitting
  train_df_full <- all_data |>
    filter(Date < train_cutoff) |>
    select(Date, all_of(c(target, feature_cols))) |>
    na.omit()

  train_df <- train_df_full |> select(-Date)

  message(model_name, " | n_train = ", nrow(train_df),
          " | p = ", length(feature_cols))

  # Define model spec
  xgb_spec <- boost_tree(
    trees      = trees,
    tree_depth = tree_depth,
    learn_rate = learn_rate,
    min_n      = min_n
  ) |>
    set_engine("xgboost") |>
    set_mode("regression")

  # Build workflow with recipe + model, then fit
  xgb_wf <- workflow() |>
    add_recipe(recipe(as.formula(paste(target, "~ .")), data = train_df)) |>
    add_model(xgb_spec)

  xgb_fit <- xgb_wf |> fit(data = train_df)

  # In-sample fitted values
  train_df_full <- train_df_full |>
    mutate(fitted = predict(xgb_fit, new_data = train_df)$.pred)

  # Training RMSE
  train_rmse <- sqrt(mean((train_df_full[[target]] - train_df_full$fitted)^2,
                           na.rm = TRUE))

  # Training plot: observed vs fitted timeseries
  p_fit <- ggplot(train_df_full) +
    geom_point(aes(x = Date, y = !!sym(target), color = "Observed"),
               size = 1.2, alpha = 0.8) +
    geom_point(aes(x = Date, y = fitted, color = "XGB fitted"),
               size = 1.2, alpha = 0.8) +
    scale_color_manual(values = c("Observed" = "black", "XGB fitted" = "dodgerblue"),
                       breaks = c("Observed", "XGB fitted")) +
    labs(title = paste0(model_name, " — training fit (pre-", year(train_cutoff), ")",
                        "  |  RMSE = ", round(train_rmse, 2), " QSU"),
         x = "Date", y = "fDOM (QSU)", color = NULL) +
    theme_bw() +
    theme(legend.position = "top")

  # Variable importance — extract raw xgboost engine and use xgb.importance directly
  # (bypasses vip/parsnip wrapper that triggers a JSON casting bug in newer xgboost)
  xgb_engine <- extract_fit_engine(xgb_fit)
  imp_df <- xgboost::xgb.importance(model = xgb_engine,
                                     feature_names = feature_cols) |>
    as_tibble() |>
    rename(Variable = Feature, Importance = Gain) |>
    arrange(desc(Importance))

  p_imp <- ggplot(imp_df, aes(x = Importance, y = reorder(Variable, Importance))) +
    geom_col(fill = "#F8766D") +
    labs(title = paste0(model_name, " — variable importance"),
         x = "Importance (XGBoost gain)", y = NULL) +
    theme_bw() +
    theme(text = element_text(size = 12))

  list(
    model          = xgb_fit,
    plot           = p_fit,
    importance_plot = p_imp,
    model_name     = model_name,
    target         = target,
    lag_col        = lag_col,
    covariates     = covariates,
    train_rmse     = train_rmse
  )
}
```

### Fit 4 models

All models trained on data before `r train_cutoff`.

```{r fit-models}
# # XGB1: lag only
# XGB_lagonly <- fit_xgb_model(
#   all_data     = ar_df,
#   train_cutoff = train_cutoff,
#   target       = "fDOM_1_QSU_daily",
#   lag_col      = "fDOM_1m_lag1",
#   covariates   = NULL,
#   model_name   = "XGB_lag_only"
# )

# XGB2: lag + in-lake (Temp, DO, Chla)
XGB_lake <- fit_xgb_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  lag_col      = "fDOM_1m_lag1",
  covariates   = c("Temp_1_C_daily", "DOsat_1_pct_daily", "Chla_1_ugL_daily"),
  model_name   = "XGB_lag_lake"
)

# XGB3: lag + catchment (RH discharge and DOC)
XGB_catch <- fit_xgb_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  lag_col      = "fDOM_1m_lag1",
  covariates   = c("RH_Q_m3day", "RH_DOC_mgL"),
  model_name   = "XGB_lag_catchment"
)

# XGB4: lag + in-lake + catchment
XGB_lake_catch <- fit_xgb_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  lag_col      = "fDOM_1m_lag1",
  covariates   = c("Temp_1_C_daily", "DOsat_1_pct_daily", "Chla_1_ugL_daily",
                   "RH_Q_m3day", "RH_DOC_mgL"),
  model_name   = "XGB_lag_lake_catchment"
)
```

### Training plots

```{r plot-training, fig.height=4}
# XGB_lagonly$plot
XGB_lake$plot
XGB_catch$plot
XGB_lake_catch$plot
```

### Variable importance plots

```{r plot-importance, fig.height=4}
# XGB_lagonly$importance_plot
XGB_lake$importance_plot
XGB_catch$importance_plot
XGB_lake_catch$importance_plot
```

---

## 30-Day Forecast Function

Same structure as the AR and RF versions. The only difference is that parsnip's
`predict()` takes `new_data` (not `newdata`) and returns a tibble — `.pred` is
extracted rather than wrapping in `as.numeric()`.

```{r fn-forecast}
# # inputs to test function
# forecast_date <- ymd("2024-01-01")
# model_obj <- XGB_lake_catch
# all_data <- ar_df
# target  = "fDOM_1_QSU_daily"
# horizon = 30

forecast_fdom_30day <- function(forecast_date,
                                 model_obj,
                                 all_data,
                                 target  = "fDOM_1_QSU_daily",
                                 horizon = 30) {

  lag_col    <- model_obj$lag_col
  covariates <- model_obj$covariates
  model      <- model_obj$model

  # Seed: extracted BEFORE masking — reference date observation is valid IC
  curr_fdom <- all_data |>
    filter(Date == forecast_date) |>
    pull(!!sym(target))

  if (length(curr_fdom) == 0 || is.na(curr_fdom)) return(NULL)

  # Mask future fDOM — prevents any observed or lagged fDOM on/after forecast_date
  # from being accessible during the forecast loop
  all_data <- all_data |>
    mutate(
      !!sym(target)  := if_else(Date >= forecast_date, NA_real_, !!sym(target)),
      !!sym(lag_col) := if_else(Date >  forecast_date, NA_real_, !!sym(lag_col))
    )

  # Sequence of dates: reference date through 30 days ahead
  forecast_dates <- seq(from = forecast_date,
                        to   = forecast_date + horizon,
                        by   = "day")

  # Pre-allocate forecast dataframe
  forecast_df <- tibble(
    model_id          = model_obj$model_name,
    reference_date    = forecast_date,
    forecast_date     = forecast_dates,
    Horizon           = as.integer(forecast_dates - reference_date),
    fDOM_1m_predicted = as.double(NA)
  )

  # Seed horizon 0 with the observed value (used as lag for step 1)
  forecast_df$fDOM_1m_predicted[1] <- curr_fdom

  # Step forward one day at a time
  for (i in 2:length(forecast_dates)) {

    # Lag: prior step's prediction pulled from the forecast dataframe
    fdom_lag <- forecast_df$fDOM_1m_predicted[i - 1]

    # Covariates: observed values for this forecast date
    fc_row <- all_data |> filter(Date == forecast_dates[i])
    if (nrow(fc_row) == 0) next

    # Assemble newdata for predict() — parsnip requires a tibble via new_data
    newdata <- setNames(tibble(fdom_lag), lag_col)
    if (!is.null(covariates)) {
      for (cv in covariates) newdata[[cv]] <- fc_row[[cv]]
    }

    # parsnip predict returns a tibble; extract .pred
    forecast_df$fDOM_1m_predicted[i] <- predict(model, new_data = newdata)$.pred
  }

  # Return only the 30 forecast days (drop horizon 0 seed row)
  forecast_df |> filter(Horizon > 0)
} # end function

# # test function
# zz <- forecast_fdom_30day(forecast_date = ymd("2024-01-01"), model_obj = XGB_lake_catch,
#                     all_data = ar_df, target = "fDOM_1_QSU_daily", horizon = 30)
```

---

## Run Forecasts Across 2024

Identical to the AR and RF versions.

```{r fn-loop}
run_forecasts <- function(model_obj,
                           all_data,
                           pred_dates,
                           target  = "fDOM_1_QSU_daily",
                           horizon = 30) {

  # Keep only dates with an observed target value (needed to seed the lag)
  valid_dates <- all_data |>
    filter(Date %in% pred_dates, !is.na(!!sym(target))) |>
    pull(Date)

  message(model_obj$model_name, " | forecasting ", length(valid_dates),
          " reference dates...")

  map_dfr(valid_dates, \(d) forecast_fdom_30day(
    forecast_date = d,
    model_obj     = model_obj,
    all_data      = all_data,
    target        = target,
    horizon       = horizon
  ))
}
```

```{r run-forecasts}
# forecasts_xgb1 <- run_forecasts(XGB_lagonly,    ar_df, pred_dates_2024)
forecasts_xgb2 <- run_forecasts(XGB_lake,       ar_df, pred_dates_2024)
forecasts_xgb3 <- run_forecasts(XGB_catch,      ar_df, pred_dates_2024)
forecasts_xgb4 <- run_forecasts(XGB_lake_catch, ar_df, pred_dates_2024)

# Combine all models into one dataframe
all_XGB_forecasts <- bind_rows(#forecasts_xgb1, 
                                forecasts_xgb2,
                                forecasts_xgb3, forecasts_xgb4)

glimpse(all_XGB_forecasts)
```

---

## Save Output

```{r save}
write_csv(all_XGB_forecasts,
          "../Data/XGB_forecasts_2024_24may26.csv")

```
