---
title: "NNETAR fDOM Forecasting — Streamlined Workflow"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5
)
```

---

## Packages and Data

Uses `forecast::nnetar()` rather than `fable::NNETAR()`. The key advantage is the
`model =` argument: passing the originally fitted model object updates the AR lag state
from new observed data **without refitting the network weights**. This gives the same
fit-once / forecast-anywhere structure as the AR, RF, and XGB scripts.

No `lag_col` argument is needed — NNETAR selects its autoregressive order `p`
automatically (minimises AIC internally).

```{r packages}
library(tidyverse)
library(lubridate)
library(forecast)   # nnetar(), forecast.nnetar()
library(zoo)        # na.approx for gap-filling
```

```{r load-data}
getwd()
ar_df <- read_csv("../Data/Daily_catwalk_met_RH_2021_2025_lagZS.csv") |>
  mutate(Date = as.Date(Date)) |>
  select(Date, fDOM_1_QSU_daily,           # no lag column needed — NNETAR handles AR internally
         Temp_1_C_daily, DOsat_1_pct_daily, Chla_1_ugL_daily,
         RH_Q_m3day, RH_DOC_mgL)

# Training window: all data before 2024
train_cutoff <- ymd("2024-01-01")

# Prediction window: every day of 2024
pred_dates_2024 <- seq(ymd("2024-01-01"), ymd("2024-12-31"), by = "day")
```

---

## Helper: Interpolate NAs

NNETAR requires a complete, gap-free numeric vector. Missing values are linearly
interpolated; leading/trailing NAs are filled by the nearest observed value (`rule = 2`).

```{r fn-interpolate}
interpolate_ts <- function(x) {
  if (all(is.na(x))) return(x)
  zoo::na.approx(x, na.rm = FALSE, rule = 2)
}
```

---

## Model Fitting

### Fitting function

Fits a NNETAR model via `forecast::nnetar()` on data before `train_cutoff`.
`repeats = 20` averages over 20 random network initialisations for stability
(Melofton et al. approach). Autoregressive order `p` and hidden node count `size`
are selected automatically.

External covariates are passed as the `xreg` matrix argument, equivalent to the
multivariate NNETAR formulation in Melofton et al.

There is no variable importance plot — NNETAR neural-network weights are not
interpretable in the way RF permutation importance or XGBoost gain are.

Returns the fitted model, training plot, and metadata needed for forecasting.

```{r fn-fit}
fit_nnetar_model <- function(all_data,
                              train_cutoff,
                              target,
                              covariates = NULL,
                              model_name,
                              repeats    = 20,
                              seed       = 100) {

  set.seed(seed)

  # Build training data — complete date sequence then interpolate NAs
  train_df <- all_data |>
    filter(Date < train_cutoff) |>
    select(Date, all_of(c(target, covariates))) |>
    arrange(Date) |>
    tidyr::complete(Date = seq(min(Date), max(Date), by = "day")) |>
    mutate(across(-Date, interpolate_ts))

  y_train    <- train_df[[target]]
  xreg_train <- if (!is.null(covariates)) as.matrix(train_df[, covariates]) else NULL

  message(model_name, " | n_train = ", nrow(train_df),
          " | covariates = ", ifelse(is.null(covariates), "none",
                                     paste(covariates, collapse = ", ")))

  # Fit NNETAR — network weights determined here and held fixed during forecasting
  mod <- nnetar(y = y_train, xreg = xreg_train, repeats = repeats)

  # In-sample fitted values (average across repeats, stored in mod$fitted)
  train_df <- train_df |>
    mutate(fitted = as.numeric(mod$fitted))

  # Training RMSE
  train_rmse <- sqrt(mean((train_df[[target]] - train_df$fitted)^2, na.rm = TRUE))

  # Training plot: observed vs fitted timeseries
  p_fit <- ggplot(train_df) +
    geom_point(aes(x = Date, y = !!sym(target), color = "Observed"),
               size = 1.2, alpha = 0.8) +
    geom_point(aes(x = Date, y = fitted, color = "NNETAR fitted"),
               size = 1.2, alpha = 0.8) +
    scale_color_manual(values = c("Observed" = "black", "NNETAR fitted" = "dodgerblue"),
                       breaks = c("Observed", "NNETAR fitted")) +
    labs(title = paste0(model_name, " — training fit (pre-", year(train_cutoff), ")",
                        "  |  RMSE = ", round(train_rmse, 2), " QSU"),
         x = "Date", y = "fDOM (QSU)", color = NULL) +
    theme_bw() +
    theme(legend.position = "top")

  list(
    model      = mod,
    plot       = p_fit,
    model_name = model_name,
    target     = target,
    covariates = covariates,
    repeats    = repeats,
    train_rmse = train_rmse
  )
}
```

### Fit 4 models

All models trained on data before `r train_cutoff`. Network weights are fixed here
and reused for all 2024 forecasts.

```{r fit-models}
# NNETAR1: univariate (AR lags of fDOM only)
NNETAR_lagonly <- fit_nnetar_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  covariates   = NULL,
  model_name   = "NNETAR_lag_only"
)

# NNETAR2: + in-lake (Temp, DO, Chla)
NNETAR_lake <- fit_nnetar_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  covariates   = c("Temp_1_C_daily", "DOsat_1_pct_daily", "Chla_1_ugL_daily"),
  model_name   = "NNETAR_lag_lake"
)

# NNETAR3: + catchment (RH discharge and DOC)
NNETAR_catch <- fit_nnetar_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  covariates   = c("RH_Q_m3day", "RH_DOC_mgL"),
  model_name   = "NNETAR_lag_catchment"
)

# NNETAR4: + in-lake + catchment
NNETAR_lake_catch <- fit_nnetar_model(
  all_data     = ar_df,
  train_cutoff = train_cutoff,
  target       = "fDOM_1_QSU_daily",
  covariates   = c("Temp_1_C_daily", "DOsat_1_pct_daily", "Chla_1_ugL_daily",
                   "RH_Q_m3day", "RH_DOC_mgL"),
  model_name   = "NNETAR_lag_lake_catchment"
)
```

### Training plots

```{r plot-training, fig.height=4}
NNETAR_lagonly$plot
NNETAR_lake$plot
NNETAR_catch$plot
NNETAR_lake_catch$plot
```

---

## 30-Day Forecast Function

The network weights are **never retrained** — the original fitted model is reused for
every reference date.

`nnetar(..., model = model_obj$model)` feeds the observed history up to the reference
date through the fixed network to update the autoregressive lag state (essentially
setting the correct input vector for the first forecast step). This is near-instantaneous
because no optimisation is performed.

`forecast::forecast()` then steps forward 30 days using those updated lag inputs,
propagating uncertainty via bootstrap sampling.

Observed covariate values for the 30 forecast days are supplied via `xreg`
(perfect covariate assumption, same as AR / RF / XGB).

```{r fn-forecast}
# # inputs to test function
forecast_date <- ymd("2024-01-01")
model_obj <- NNETAR_lake_catch
all_data <- ar_df
target  = "fDOM_1_QSU_daily"
horizon = 30
bootstrap_times = 200
seed            = 100

forecast_fdom_30day <- function(forecast_date,
                                 model_obj,
                                 all_data,
                                 target          = "fDOM_1_QSU_daily",
                                 horizon         = 30,
                                 seed            = 100) {

  set.seed(seed)

  covariates <- model_obj$covariates

  # Check that observed fDOM exists on the reference date
  curr_fdom <- all_data |>
    filter(Date == forecast_date) |>
    pull(!!sym(target))

  if (length(curr_fdom) == 0 || is.na(curr_fdom)) return(NULL)

  # History up to and including the reference date — used to update the AR lag state
  history_df <- all_data |>
    filter(Date <= forecast_date) |>
    select(Date, all_of(c(target, covariates))) |>
    arrange(Date) |>
    tidyr::complete(Date = seq(min(Date), max(Date), by = "day")) |>
    mutate(across(-Date, interpolate_ts))

  y_history    <- history_df[[target]]
  xreg_history <- if (!is.null(covariates)) as.matrix(history_df[, covariates]) else NULL

  # Update AR lag state using fixed network weights (no retraining)
  mod_updated <- nnetar(y = y_history, xreg = xreg_history,
                         model = model_obj$model)

  # Forecast dates: day 1 through day 30 after the reference date
  forecast_dates <- seq(forecast_date + 1, forecast_date + horizon, by = "day")

  # Covariate values for the 30 forecast days (perfect covariate assumption)
  if (!is.null(covariates)) {
    future_df <- all_data |>
      filter(Date %in% forecast_dates) |>
      select(Date, all_of(covariates)) |>
      arrange(Date) |>
      tidyr::complete(Date = forecast_dates) |>
      mutate(across(-Date, interpolate_ts))

    xreg_future <- as.matrix(future_df[, covariates])
    if (nrow(xreg_future) == 0) return(NULL)
  } else {
    xreg_future <- NULL
  }

  # Generate 30-day forecast with bootstrap uncertainty
  fc <- forecast(mod_updated, h = horizon, xreg = xreg_future)

  # Return mean prediction — format matches AR / RF / XGB output
  tibble(
    model_id          = model_obj$model_name,
    reference_date    = forecast_date,
    forecast_date     = forecast_dates,
    Horizon           = as.integer(forecast_dates - reference_date),
    fDOM_1m_predicted = as.numeric(fc$mean)
  )
}

# # test function
zz <- forecast_fdom_30day(forecast_date = ymd("2024-01-01"), model_obj = NNETAR_lake_catch,
                    all_data = ar_df, target = "fDOM_1_QSU_daily", horizon = 30)

```

---

## Run Forecasts Across 2024

Identical structure to the AR, RF, and XGB versions.

```{r fn-loop}
run_forecasts <- function(model_obj,
                           all_data,
                           pred_dates,
                           target  = "fDOM_1_QSU_daily",
                           horizon = 30) {

  # Keep only dates with an observed target value (needed to confirm valid seed)
  valid_dates <- all_data |>
    filter(Date %in% pred_dates, !is.na(!!sym(target))) |>
    pull(Date)

  message(model_obj$model_name, " | forecasting ", length(valid_dates),
          " reference dates...")

  map_dfr(valid_dates, \(d) forecast_fdom_30day(
    forecast_date = d,
    model_obj     = model_obj,
    all_data      = all_data,
    target        = target,
    horizon       = horizon
  ))
}
```

```{r run-forecasts}
forecasts_nnetar1 <- run_forecasts(NNETAR_lagonly,    ar_df, pred_dates_2024)
forecasts_nnetar2 <- run_forecasts(NNETAR_lake,       ar_df, pred_dates_2024)
forecasts_nnetar3 <- run_forecasts(NNETAR_catch,      ar_df, pred_dates_2024)
forecasts_nnetar4 <- run_forecasts(NNETAR_lake_catch, ar_df, pred_dates_2024)

# Combine all models into one dataframe
all_NNETAR_forecasts <- bind_rows(forecasts_nnetar1, forecasts_nnetar2,
                                   forecasts_nnetar3, forecasts_nnetar4)

glimpse(all_NNETAR_forecasts)
```

---

## Save Output

```{r save}
write_csv(all_NNETAR_forecasts,
          "../Data/NNETAR_forecasts_2024_22may26.csv")

```
