---
title: "AR Model Forecasting — fDOM 1m (30-day ahead)"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5
)
```

This document fits a suite of AR (autoregressive) GLMs to predict fDOM at 1m depth,
generates iterative 30-day ahead forecasts for all of 2024, and evaluates skill by
forecast horizon and season.

Each model is defined by a target variable, a lag column, and optional covariates.
All model fitting, forecasting, plotting, and RMSE calculation is handled by four
reusable functions defined in the **Function Definitions** section below.

**Forecasting approach** (following Lofton et al.): observed covariates are used at
each forecast timestep; the lag term is updated iteratively from the prior day's
prediction rather than from observed values.

---

## Packages and Data

```{r packages}
library(tidyverse)
library(lubridate)
```

```{r load-data}
ar_df <- read_csv("./CCR_AR_Modeling/Data/Daily_catwalk_met_RH_2021_2025_lagZS.csv") |>
  mutate(
    Date = as.Date(Date),
    RH_DOCload_10day = slider::slide_dbl(RH_streamflow_DOC_gm2day, mean,
                                         .before = 10, .after = 0, .complete = F),
    RH_Q_7daymean   = slider::slide_dbl(RH_Q_m3day,   mean,
                                         .before = 7,  .after = 0, .complete = F),
    RH_DOC_7daymean = slider::slide_dbl(RH_DOC_mgL,   mean,
                                         .before = 7,  .after = 0, .complete = F),
    Chla_7daymean   = slider::slide_dbl(Chla_1_ugL_daily, mean,
                                         .before = 7,  .after = 0, .complete = F)
  )

# All 2024 calendar dates — fit_ar_model filters to those with observed target
pred_dates_2024 <- ar_df |>
  filter(Date >= ymd("2024-01-01"), Date < ymd("2025-01-01")) |>
  pull(Date)
```

---

## Function Definitions

### Internal: iterative forecast engine

Called inside `fit_ar_model()`; not usually used directly. Steps forward one day at
a time, using the prior day's prediction as the lag and observed covariate values
for each forecast date.

```{r fn-forecast-engine}
forecast_ar_gen <- function(ref_date, model, model_name, horizon = 30,
                             all_data, target, lag_col, covar_cols = NULL) {

  obs_ref <- all_data |> filter(Date == ref_date) |> pull(!!sym(target))
  if (length(obs_ref) == 0 || is.na(obs_ref)) return(NULL)

  prev_pred <- obs_ref
  results   <- vector("list", horizon)

  for (h in seq_len(horizon)) {
    forecast_date <- ref_date + h
    fc_row <- all_data |> filter(Date == forecast_date)
    if (nrow(fc_row) == 0) next

    newdata <- setNames(data.frame(prev_pred), lag_col)
    if (!is.null(covar_cols)) {
      for (cv in covar_cols) newdata[[cv]] <- fc_row[[cv]]
    }

    pred_val <- as.numeric(predict(model, newdata = newdata))

    results[[h]] <- tibble(
      model_id       = model_name,
      reference_date = ref_date,
      forecast_date  = forecast_date,
      horizon        = h,
      prediction     = pred_val
    )
    prev_pred <- pred_val
  }

  bind_rows(results)
}
```

### Main wrapper: `fit_ar_model()`

Fits a GLM, runs iterative 30-day forecasts for all valid 2024 reference dates,
and returns a named list containing the model, training data with fitted values,
and forecast output with observed values joined.

```{r fn-fit-model}
fit_ar_model <- function(all_data,
                          train_cutoff = ymd("2024-01-01"),
                          pred_dates,
                          target,
                          lag_col,
                          covariates = NULL,
                          model_name,
                          horizon = 30) {

  # Build training data: filter to pre-cutoff, drop NAs in all model variables
  train_df <- all_data |>
    filter(Date < train_cutoff,
           !is.na(!!sym(target)),
           !is.na(!!sym(lag_col)))

  if (!is.null(covariates)) {
    for (cv in covariates) train_df <- train_df |> filter(!is.na(!!sym(cv)))
  }

  # Build formula and fit GLM
  rhs     <- paste(c(lag_col, covariates), collapse = " + ")
  formula <- as.formula(paste(target, "~", rhs))
  mod     <- glm(formula, data = train_df, family = gaussian, na.action = na.fail)
  print(summary(mod))

  # Add in-sample fitted values to training data
  train_fitted <- train_df |> mutate(fitted = predict(mod, newdata = train_df))

  # Filter pred_dates to rows with observed target (needed to seed the lag)
  valid_pred_dates <- all_data |>
    filter(Date %in% pred_dates, !is.na(!!sym(target))) |>
    pull(Date)

  message(model_name, " | n_train = ", nrow(train_df),
          " | n_ref_dates = ", length(valid_pred_dates))

  # Run iterative forecasts
  forecasts <- map_dfr(valid_pred_dates, \(d) forecast_ar_gen(
    ref_date   = d, model = mod, model_name = model_name,
    horizon    = horizon, all_data = all_data,
    target     = target, lag_col = lag_col, covar_cols = covariates
  ))

  # Join observed target onto forecasts by forecast_date
  forecasts_obs <- forecasts |>
    left_join(all_data |> select(Date, fDOM_obs = !!sym(target)),
              by = c("forecast_date" = "Date"))

  list(model      = mod,      model_name = model_name,
       target     = target,   lag_col    = lag_col,
       covariates = covariates,
       train_data = train_fitted,
       forecasts  = forecasts_obs)
}
```

### Plot: `plot_ar_forecast()`

Produces a single timeseries showing observed (black), training fit (orange), and
test-period predictions (blue) for a chosen forecast horizon.

```{r fn-plot}
plot_ar_forecast <- function(model_obj, horizon,
                              train_cutoff = as.Date("2024-01-01")) {

  target     <- model_obj$target
  model_name <- model_obj$model_name
  df_train   <- model_obj$train_data
  df_test    <- model_obj$forecasts |> filter(horizon == !!horizon)

  obs_data <- bind_rows(
    df_train |> select(Date, fDOM = !!sym(target)),
    df_test  |> select(Date = forecast_date, fDOM = fDOM_obs)
  ) |> filter(!is.na(fDOM)) |> mutate(series = "Observed")

  train_fit <- df_train |>
    filter(!is.na(fitted)) |>
    select(Date, fDOM = fitted) |>
    mutate(series = "Training fit")

  test_pred <- df_test |>
    filter(!is.na(prediction)) |>
    select(Date = forecast_date, fDOM = prediction) |>
    mutate(series = "Test prediction")

  bind_rows(obs_data, train_fit, test_pred) |>
    ggplot(aes(x = Date, y = fDOM, color = series)) +
    geom_point(size = 1.2, alpha = 0.8) +
    geom_vline(xintercept = train_cutoff, linetype = "dashed", color = "gray40") +
    scale_color_manual(
      values = c("Observed" = "black", "Training fit" = "orange",
                 "Test prediction" = "dodgerblue"),
      breaks = c("Observed", "Training fit", "Test prediction")
    ) +
    labs(title = paste0(model_name, "  |  horizon = ", horizon, " day(s)"),
         x = "Date", y = "fDOM (QSU)", color = NULL) +
    theme_bw() + theme(legend.position = "top")
}
```

### RMSE: `calc_rmse_by_horizon()`

Returns RMSE at each forecast horizon. Set `by_season = TRUE` to additionally
group by season (Winter / Spring / Summer / Fall).

```{r fn-rmse}
calc_rmse_by_horizon <- function(model_obj, by_season = FALSE) {

  df <- model_obj$forecasts |>
    filter(!is.na(prediction), !is.na(fDOM_obs))

  if (by_season) {
    df <- df |>
      mutate(
        month  = month(forecast_date),
        Season = case_when(
          month %in% c(12, 1, 2)  ~ "Winter",
          month %in% c(3,  4, 5)  ~ "Spring",
          month %in% c(6,  7, 8)  ~ "Summer",
          month %in% c(9, 10, 11) ~ "Fall"
        )
      )
    df |>
      group_by(model_id, horizon, Season) |>
      summarise(RMSE = sqrt(mean((prediction - fDOM_obs)^2)), .groups = "drop")
  } else {
    df |>
      group_by(model_id, horizon) |>
      summarise(RMSE = sqrt(mean((prediction - fDOM_obs)^2)), .groups = "drop")
  }
}
```

---

## Run Models

Models are grouped by covariate set. All use `fDOM_1m_lag1` as the lag predictor
and `fDOM_1_QSU_daily` as the target. Summaries print to confirm coefficient
estimates and significance.

```{r run-lag-only, results='hide'}
## M1 — lag only
m1_1m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_1_QSU_daily", lag_col = "fDOM_1m_lag1",
  covariates = NULL, model_name = "M1_1m_lag"
)

m1_9m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_9_QSU_daily", lag_col = "fDOM_9m_lag1",
  covariates = NULL, model_name = "M1_9m_lag"
)

```

```{r run-lake-models, results='hide'}
## M2a — lag + Temp + DO + SW
m2_1m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_1_QSU_daily", lag_col = "fDOM_1m_lag1",
  covariates = c("Temp_1_C_daily", "DOsat_1_pct_daily", "SW_Wm2_daily"),
  model_name = "M2_1m_lag_LAKEa"
)

## M2b — lag + Temp + DO
m2b_1m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_1_QSU_daily", lag_col = "fDOM_1m_lag1",
  covariates = c("Temp_1_C_daily", "DOsat_1_pct_daily"),
  model_name = "M2_1m_lag_LAKEb"
)

m2b_9m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_9_QSU_daily", lag_col = "fDOM_9m_lag1",
  covariates = c("Temp_9_C_daily", "DOsat_9_pct_daily"),
  model_name = "M2_9m_lag_LAKEb"
)

## M2c — lag + Temp + DO + Chla 7-day mean
m2c_1m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_1_QSU_daily", lag_col = "fDOM_1m_lag1",
  covariates = c("Temp_1_C_daily", "DOsat_1_pct_daily", "Chla_7daymean"),
  model_name = "M2_1m_lag_LAKEc"
)
```

```{r run-catchment-models, results='hide'}
## M4a — lag + Temp + DO + SW + RH Q + RH DOC
m4_1m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_1_QSU_daily", lag_col = "fDOM_1m_lag1",
  covariates = c("Temp_1_C_daily", "DOsat_1_pct_daily", "SW_Wm2_daily",
                 "RH_Q_m3day", "RH_DOC_mgL"),
  model_name = "M4_1m_lag_LAKEa_RH"
)

## M4b — lag + Temp + DO + RH Q + RH DOC
m4b_1m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_1_QSU_daily", lag_col = "fDOM_1m_lag1",
  covariates = c("Temp_1_C_daily", "DOsat_1_pct_daily",
                 "RH_Q_m3day", "RH_DOC_mgL"),
  model_name = "M4_1m_lag_LAKEb_RH"
)

m4b_9m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_9_QSU_daily", lag_col = "fDOM_9m_lag1",
  covariates = c("Temp_9_C_daily", "DOsat_9_pct_daily",
                 "RH_Q_m3day", "RH_DOC_mgL"),
  model_name = "M4_9m_lag_LAKEb_RH"
)


## M5a — lag + RH Q + RH DOC (daily)
m5_1m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_1_QSU_daily", lag_col = "fDOM_1m_lag1",
  covariates = c("RH_Q_m3day", "RH_DOC_mgL"),
  model_name = "M5_1m_lag_RH"
)

## M5a — lag + RH Q + RH DOC (daily)
m5_9m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_9_QSU_daily", lag_col = "fDOM_9m_lag1",
  covariates = c("RH_Q_m3day", "RH_DOC_mgL"),
  model_name = "M5_9m_lag_RH"
)

## M5b — lag + RH Q + RH DOC (7-day means)
m5b_1m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_1_QSU_daily", lag_col = "fDOM_1m_lag1",
  covariates = c("RH_Q_7daymean", "RH_DOC_7daymean"),
  model_name = "M5_1m_lag_RHlag"
)
```

```{r run-rain-models, results='hide'}
## M6a — lag + Temp + DO + SW + Rain lag1
m6_1m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_1_QSU_daily", lag_col = "fDOM_1m_lag1",
  covariates = c("Temp_1_C_daily", "DOsat_1_pct_daily", "SW_Wm2_daily",
                 "Rain_mm_lag1"),
  model_name = "M6_1m_lag_Lakea_Rain"
)

## M6b — lag + Temp + DO + Rain lag1
m6b_1m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_1_QSU_daily", lag_col = "fDOM_1m_lag1",
  covariates = c("Temp_1_C_daily", "DOsat_1_pct_daily", "Rain_mm_lag1"),
  model_name = "M6_1m_lag_Lakeb_Rain"
)

## M7 — lag + Rain lag1 only
m7_1m <- fit_ar_model(
  all_data = ar_df, pred_dates = pred_dates_2024,
  target = "fDOM_1_QSU_daily", lag_col = "fDOM_1m_lag1",
  covariates = c("Rain_mm_lag1"),
  model_name = "M7_1m_lag_Rain"
)
```

---

## Collect Outputs and Timeseries Plots

### Combine all forecasts

```{r collect-forecasts}
model_list <- list(m1_1m, m2b_1m, #m2_1m, m2c_1m, m4_1m, m6_1m, m5b_1m # SW or chla
                   m4b_1m, m5_1m #, m6b_1m, m7_1m #Rains
                   )

all_forecasts_obs <- map_dfr(model_list, \(m) m$forecasts)

unique(all_forecasts_obs$model_id)
```

### Training vs test plots

Black points = observed. Orange = in-sample training fit. Blue = out-of-sample
test prediction at the chosen horizon.

```{r train-test-plots}
plot_ar_forecast(m1_1m, horizon = 1)
plot_ar_forecast(m1_1m, horizon = 30)
plot_ar_forecast(m2_1m, horizon = 7)
plot_ar_forecast(m4_1m, horizon = 7)
```

### Multi-horizon timeseries (one model)

```{r ts-multi-horizon}
all_forecasts_obs |>
  filter(horizon %in% c(1, 7, 15, 30),
         model_id == "M5_1m_lag_RH") |>
  ggplot() +
  geom_line(aes(x = forecast_date, y = prediction,
                col = as.factor(horizon)), linewidth = 1.1) +
  geom_point(aes(x = forecast_date, y = fDOM_obs), size = 0.8) +
  labs(x = "Date", y = "fDOM (QSU)", col = "Horizon (days)",
       title = "M5_1m_lag_RH — predictions by horizon") +
  theme_bw() + theme(legend.position = "top")
```

### Multi-model timeseries (one horizon)

```{r ts-multi-model}
all_forecasts_obs |>
  filter(horizon == 7,
         model_id %in% c("M1_1m_lag", "M2_1m_lag_LAKEa",
                          "M4_1m_lag_LAKEa_RH", "M6_1m_lag_Lakeb_Rain")) |>
  ggplot() +
  geom_line(aes(x = forecast_date, y = prediction,
                col = model_id), linewidth = 1.1) +
  geom_point(aes(x = forecast_date, y = fDOM_obs), size = 0.8) +
  labs(x = "Date", y = "fDOM (QSU)", col = "Model",
       title = "7-day ahead forecast — all models") +
  theme_bw() + theme(legend.position = "top")
```

---

## RMSE Calculations

### Overall RMSE by horizon

```{r rmse-overall}
# Shared label map — used in all RMSE chunks and the example forecast
model_labels <- c(
  "M1_1m_lag"          = "AR only",
  "M2_1m_lag_LAKEb"    = "AR + in-lake",
  "M4_1m_lag_LAKEb_RH" = "AR + in-lake + catchment",
  "M5_1m_lag_RH"       = "AR + catchment"
)
models_keep <- names(model_labels)

rmse_all <- bind_rows(
  calc_rmse_by_horizon(m1_1m),
  calc_rmse_by_horizon(m2b_1m),
  calc_rmse_by_horizon(m4b_1m),
  calc_rmse_by_horizon(m5_1m)
) |>
  mutate(model_id = recode(model_id, !!!model_labels))

ggplot(rmse_all, aes(x = horizon, y = RMSE, col = model_id)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  labs(#title = "Prediction skill by horizon",
       x = "Forecast horizon (days ahead)", y = "RMSE (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 24))
```

### Seasonal RMSE by horizon

```{r rmse-seasonal}
rmse_all_season <- bind_rows(
  calc_rmse_by_horizon(m1_1m,  by_season = TRUE),
  calc_rmse_by_horizon(m2b_1m, by_season = TRUE),
  calc_rmse_by_horizon(m4b_1m, by_season = TRUE),
  calc_rmse_by_horizon(m5_1m,  by_season = TRUE)
) |>
  mutate(model_id = recode(model_id, !!!model_labels))

ggplot(rmse_all_season, aes(x = horizon, y = RMSE, col = model_id)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~Season, nrow = 1) +
  labs(title = "Prediction skill by horizon and season",
       x = "Forecast horizon (days ahead)", y = "RMSE (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))
```

## 9m RMSEs

```{r}
# Shared label map — used in all RMSE chunks and the example forecast
model_labels <- c(
  "M1_9m_lag"          = "AR only",
  "M2_9m_lag_LAKEb"    = "AR + in-lake",
  "M4_9m_lag_LAKEb_RH" = "AR + in-lake + catchment",
  "M5_9m_lag_RH"       = "AR + catchment"
)

models_keep <- names(model_labels)

rmse_all9m <- bind_rows(
  calc_rmse_by_horizon(m1_9m),
  calc_rmse_by_horizon(m2b_9m),
  calc_rmse_by_horizon(m4b_9m),
  calc_rmse_by_horizon(m5_9m)
) |>
  mutate(model_id = recode(model_id, !!!model_labels))

ggplot(rmse_all9m, aes(x = horizon, y = RMSE, col = model_id)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  labs(title = "Prediction skill by horizon",
       x = "Forecast horizon (days ahead)", y = "RMSE (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))



##season
rmse_all_season9m <- bind_rows(
  calc_rmse_by_horizon(m1_9m,  by_season = TRUE),
  calc_rmse_by_horizon(m2b_9m, by_season = TRUE),
  calc_rmse_by_horizon(m4b_9m, by_season = TRUE),
  calc_rmse_by_horizon(m5_9m,  by_season = TRUE)
) |>
  mutate(model_id = recode(model_id, !!!model_labels))

ggplot(rmse_all_season9m, aes(x = horizon, y = RMSE, col = model_id)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~Season, nrow = 1) +
  labs(title = "Prediction skill by horizon and season",
       x = "Forecast horizon (days ahead)", y = "RMSE (QSU)", col = NULL) +
  theme_bw() + theme(legend.position = "top", text = element_text(size = 14))



```


## RMSE by Discharge Condition

Reference dates are split into low-flow (5th–15th percentile of 2024 discharge) and
high-flow (85th–95th percentile) groups. Each band captures roughly 10% of 2024 reference
dates (~30 days), providing a balanced comparison of forecast skill under contrasting
hydrological conditions.

```{r rmse-flow-thresholds}
# Compute percentile thresholds from the 2024 reference date discharge distribution
q_2024 <- all_forecasts_obs |>
  distinct(reference_date) |>
  left_join(ar_df |> select(Date, RH_Q_m3day), by = c("reference_date" = "Date"))

thresholds <- quantile(q_2024$RH_Q_m3day, probs = c(0.05, 0.15, 0.85, 0.95), na.rm = TRUE)

low_dates  <- q_2024 |> filter(RH_Q_m3day >= thresholds[1], RH_Q_m3day <= thresholds[2]) |> pull(reference_date)
high_dates <- q_2024 |> filter(RH_Q_m3day >= thresholds[3], RH_Q_m3day <= thresholds[4]) |> pull(reference_date)

print(low_dates)
print(high_dates)


# message("Low-flow reference dates (5th–15th pct):  n = ", length(low_dates),
#         "  |  Q range: ", round(thresholds[1], 2), "–", round(thresholds[2], 2), " m3/day")
# message("High-flow reference dates (85th–95th pct): n = ", length(high_dates),
#         "  |  Q range: ", round(thresholds[3], 2), "–", round(thresholds[4], 2), " m3/day")

```

```{r rmse-flow-calc}
# RMSE by horizon for each flow condition
calc_rmse_flow <- function(forecasts_df, ref_dates, flow_label) {
  forecasts_df |>
    filter(model_id %in% models_keep,
           reference_date %in% ref_dates,
           !is.na(prediction), !is.na(fDOM_obs)) |>
    group_by(model_id, horizon) |>
    summarise(RMSE = sqrt(mean((prediction - fDOM_obs)^2)), .groups = "drop") |>
    mutate(flow_condition = flow_label,
           model_id = recode(model_id, !!!model_labels))
}

rmse_low  <- calc_rmse_flow(all_forecasts_obs, low_dates,  "Low flow (5th–15th pct)")
rmse_high <- calc_rmse_flow(all_forecasts_obs, high_dates, "High flow (85th–95th pct)")

rmse_flow <- bind_rows(rmse_low, rmse_high)


### plot
ggplot(rmse_flow, aes(x = horizon, y = RMSE, col = model_id)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~flow_condition, nrow = 1) +
  labs(
    title = "Prediction skill by horizon — low vs. high discharge conditions",
    x     = "Forecast horizon (days ahead)",
    y     = "RMSE (QSU)",
    col   = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 14))
```

---

### Wettest vs. driest month

Identify which 2024 calendar month had the highest and lowest mean discharge across
reference dates, then compute RMSE by horizon for forecasts issued in each of those months.

```{r rmse-monthly-flow}
# Mean discharge by calendar month across 2024 reference dates
monthly_q <- q_2024 |>
  mutate(month = month(reference_date, label = TRUE, abbr = FALSE)) |>
  group_by(month) |>
  summarise(mean_Q = mean(RH_Q_m3day, na.rm = TRUE), .groups = "drop") |>
  arrange(mean_Q)

wettest_month <- monthly_q |> slice_max(mean_Q, n = 1) |> pull(month)
driest_month  <- monthly_q |> slice_min(mean_Q, n = 1) |> pull(month)

message("Wettest month: ", wettest_month,
        "  |  mean Q = ", round(monthly_q$mean_Q[monthly_q$month == wettest_month], 2), " m3/day")
message("Driest month:  ", driest_month,
        "  |  mean Q = ", round(monthly_q$mean_Q[monthly_q$month == driest_month], 2), " m3/day")

wet_dates <- q_2024 |>
  mutate(month = month(reference_date, label = TRUE, abbr = FALSE)) |>
  filter(month == wettest_month) |>
  pull(reference_date)

dry_dates <- q_2024 |>
  mutate(month = month(reference_date, label = TRUE, abbr = FALSE)) |>
  filter(month == driest_month) |>
  pull(reference_date)

rmse_wet <- calc_rmse_flow(all_forecasts_obs, wet_dates,
                            paste0("Wettest month (", wettest_month, ")"))
rmse_dry <- calc_rmse_flow(all_forecasts_obs, dry_dates,
                            paste0("Driest month (", driest_month, ")"))

rmse_monthly_flow <- bind_rows(rmse_wet, rmse_dry)
```

```{r rmse-monthly-flow-plot, fig.height = 6}
ggplot(rmse_monthly_flow, aes(x = horizon, y = RMSE, col = model_id)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~flow_condition, nrow = 1) +
  labs(
    title = "Prediction skill by horizon — wettest vs. driest month",
    x     = "Forecast horizon (days ahead)",
    y     = "RMSE (QSU)",
    col   = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 14))
```

---

## Model Skill Relative to Persistence

Skill scores express improvement over the "AR only" model (treated as persistence here).
A skill of 0 means no improvement; positive values indicate better-than-persistence skill;
negative values indicate worse performance. Calculated as:

**Skill = 1 − (model RMSE / AR-only RMSE)**

```{r skill-calc}
skill_rmse <- all_forecasts_obs |>
  filter(model_id %in% models_keep,
         !is.na(prediction), !is.na(fDOM_obs)) |>
  mutate(model_id = recode(model_id, !!!model_labels)) |>
  mutate(month  = month(forecast_date),
        Season = case_when(
          month %in% c(12, 1, 2)  ~ "Winter",
          month %in% c(3,  4, 5)  ~ "Spring",
          month %in% c(6,  7, 8)  ~ "Summer",
          month %in% c(9, 10, 11) ~ "Fall")) |> 
  group_by(model_id, horizon, Season) |>
  summarise(RMSE = sqrt(mean((prediction - fDOM_obs)^2)), .groups = "drop") |>
  pivot_wider(names_from = model_id, values_from = RMSE) |>
  mutate(
    `AR + in-lake`             = 1 - (`AR + in-lake`             / `AR only`),
    `AR + catchment`           = 1 - (`AR + catchment`           / `AR only`),
    `AR + in-lake + catchment` = 1 - (`AR + in-lake + catchment` / `AR only`)
  ) |>
  # select(horizon, `AR + in-lake`, `AR + catchment`, `AR + in-lake + catchment`) |>
  # pivot_longer(-horizon, names_to = "model_id", values_to = "skill")
    select(Season, horizon, `AR + in-lake`, `AR + catchment`, `AR + in-lake + catchment`) |>
  pivot_longer(-c(Season, horizon), names_to = "model_id", values_to = "skill")
```

```{r skill-plot}
ggplot(skill_rmse, aes(x = horizon, y = skill, col = model_id)) +
  geom_hline(yintercept = 0, linewidth = 1, linetype = "dashed", color = "gray40") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  facet_wrap(~Season, nrow = 1)+
  scale_color_manual(values = c(
    "AR + in-lake"             = "#00BA38",
    "AR + in-lake + catchment" = "#619CFF",
    "AR + catchment"           = "#F8766D"
  )) +
  labs(
    #title = "Forecast skill relative to AR-only (persistence)",
    x     = "Forecast horizon (days ahead)",
    y     = "Skill score",
    col   = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top", text = element_text(size = 20))

```

---

## Example Forecast for Talk

Three-panel build showing a 30-day forecast issued on 2024-09-01. Each panel adds
a layer: (1) pre-forecast observations only, (2) model predictions, (3) observed
values during the forecast horizon to evaluate skill.

```{r talk-forecast-data}
ref_date <- as.Date("2024-09-01")

# 3 observations before ref_date + the ref_date itself (seeds the forecast)
pre_obs <- ar_df |>
  filter(Date >= ref_date - 3, Date <= ref_date) |>
  select(Date, fDOM = fDOM_1_QSU_daily) |>
  filter(!is.na(fDOM))

# Observed fDOM during the 30-day forecast window
post_obs <- ar_df |>
  filter(Date > ref_date, Date <= ref_date + 30) |>
  select(Date, fDOM = fDOM_1_QSU_daily) |>
  filter(!is.na(fDOM))

# Model predictions for this reference date — reuse shared label map
talk_fc <- all_forecasts_obs |>
  filter(reference_date == ref_date, model_id %in% models_keep) |>
  select(Date = forecast_date, fDOM = prediction, model_id) |>
  filter(!is.na(fDOM)) |>
  mutate(model_id = recode(model_id, !!!model_labels))
```

```{r talk-forecast-plots, fig.width = 6, fig.height = 5}
x_lim        <- c(ref_date - 3, ref_date + 30)
y_lim        <- c(2, 20)
model_colors <- c("AR only"                  = "#F8766D",
                  "AR + in-lake"             = "#00BA38",
                  "AR + in-lake + catchment" = "#619CFF",
                  "AR + catchment"           = "#C77CFF")

base_layers <- list(
  geom_vline(xintercept = ref_date, linetype = "dashed", color = "gray40"),
  scale_x_date(limits = x_lim, date_labels = "%b %d"),
  scale_y_continuous(limits = y_lim),
  labs(x = "Date", y = "fDOM (QSU)", color = "Model",
       title = "Example forecast for talk"),
  theme_bw(),
  theme(legend.position = "top", text = element_text(size = 13))
)

# --- Panel 1: pre-forecast observations only ---
ggplot() +
  base_layers +
  geom_point(data = pre_obs, aes(x = Date, y = fDOM),
             color = "black", size = 2.5)

# --- Panel 2: add model forecasts (all three models) ---
ggplot() +
  base_layers +
  geom_point(data = pre_obs, aes(x = Date, y = fDOM),
             color = "black", size = 2.5) +
  geom_line(data  = talk_fc, aes(x = Date, y = fDOM, color = model_id),
            linewidth = 0.8) +
  geom_point(data = talk_fc, aes(x = Date, y = fDOM, color = model_id),
             size = 1.8) +
  scale_color_manual(values = model_colors)

# --- Panel 3: add observed values during forecast period ---
ggplot() +
  base_layers +
  geom_point(data = pre_obs,  aes(x = Date, y = fDOM),
             color = "black", size = 2.5) +
  geom_line(data  = talk_fc,  aes(x = Date, y = fDOM, color = model_id),
            linewidth = 0.8) +
  geom_point(data = talk_fc,  aes(x = Date, y = fDOM, color = model_id),
             size = 1.8) +
  geom_point(data = post_obs, aes(x = Date, y = fDOM),
             color = "black", size = 2.5) +
  scale_color_manual(values = model_colors)
```
